---
title: "Barcharts and colours"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Barcharts and colours}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
# Show each figure exactly as save_plot() writes it, at its automatic size
show_plot <- function(plot, ...){
  path.s <- knitr::fig_path(".png")
  gmaio::save_plot(plot, path.s, dpi = 110, ...)
  knitr::include_graphics(path.s, dpi = 110)
}
```

`plot_stacked_barchart()` draws the composition of each sample as a stacked bar: the most abundant taxa in colour, the rest summed into a grey `Other`.
This vignette follows `code/barcharts.R`, goes through the options for changing the figure, and then explains where the colours come from and how to change them.
It uses the synthetic example project; see `vignette("gmaio", package = "gmaio")` for how projects and profiles work.

```{r project}
project.l <- gmaio::example_project()
metadata.df <- gmaio::analysis_metadata(project.l)
```

## Choosing the data

A barchart shows one profile at one rank.
`code/barcharts.R` loops over the taxonomic profiles and the ranks in `analysis: ranks` of `config.yml`, aggregating each profile to the rank first:

```{r data}
genus.p <- gmaio::get_profile(project.l, "sylph_taxonomic") |>
  gmaio::aggregate_profile(rank = "genus")
genus.p
```

The profile is not filtered first: taxa that are not shown still count towards `Other`, so every bar adds up to 100%.

## The template barchart

```{r template}
barchart.gg <- gmaio::plot_stacked_barchart(genus.p, metadata.df, project.l$palettes$taxa,
                                            top_n = 8,
                                            facet_variable = "Treatment",
                                            legend_title = "Genus")
```

```{r template-show, echo = FALSE}
show_plot(barchart.gg)
```

Samples are drawn in metadata order (`sample_order` in `config.yml`) and labelled by `Sample_label` (`label_column`).
`facet_variable` splits the samples into panels; panel widths follow the number of samples in each.
The legend lists taxa by mean abundance, with `Unassigned` and `Other` last.

In a project, edit the `gmaio::plot_stacked_barchart()` call in `code/barcharts.R` and rerun the script.

## Which taxa are shown

By default a taxon is shown when it is among the `top_n` most abundant in **any** sample (`top_method = "per_sample"`).
Every sample then shows its own dominant taxa, but when samples differ the legend can list many more than `top_n` taxa: above, `top_n = 8` gave `r sum(!gmaio::collapse_top_n(genus.p, top_n = 8, merge_unassigned = TRUE)$features$Label %in% c("Other", "Unassigned"))` genera.
`top_method = "mean"` shows exactly the `top_n` taxa with the highest mean, which keeps the legend short but can hide a taxon that dominates only a few samples:

```{r mean}
barchart.gg <- gmaio::plot_stacked_barchart(genus.p, metadata.df, project.l$palettes$taxa,
                                            top_n = 12, top_method = "mean",
                                            facet_variable = "Treatment", legend_title = "Genus")
```

```{r mean-show, echo = FALSE}
show_plot(barchart.gg)
```

`min_abundance` additionally requires a shown taxon to reach that value (in %) in the sample where it is counted, so `top_n = 15, min_abundance = 1` shows up to 15 taxa per sample but only those at 1% or more.

Taxa with nothing assigned below domain (`Unassigned`, `Unclassified d__Bacteria`) are merged into one grey `Unassigned` segment; `merge_unassigned = FALSE` keeps them separate.
`collapse_top_n()` does the selection, so you can check what a barchart will show:

```{r collapse}
shown.p <- gmaio::collapse_top_n(genus.p, top_n = 5, method = "mean", merge_unassigned = TRUE)
round(rowMeans(shown.p$values), 1) |> stats::setNames(shown.p$features$Label)
```

## Annotation strips

`annotation_variables` adds a colour strip per metadata column under the bars, which helps when samples are split by one variable but vary in others.
Strip colours come from `annotation_colours`, usually the project palettes; `annotation_height` sets the height of each strip in cm.

```{r strips}
barchart.gg <- gmaio::plot_stacked_barchart(genus.p, metadata.df, project.l$palettes$taxa,
                                            top_n = 12, top_method = "mean",
                                            facet_variable = "Treatment",
                                            annotation_variables = c("Time", "Cage"),
                                            annotation_colours = project.l$palettes,
                                            legend_title = "Genus")
```

```{r strips-show, echo = FALSE}
show_plot(barchart.gg)
```

With annotation strips the result is a patchwork (two plots stacked) rather than a single ggplot.

## Legend, bars and text

| Argument | Controls | Default |
|---|---|---|
| `legend_title` | Legend title | none |
| `legend_ncol` | Legend columns | 1, or 2 above 25 entries |
| `legend_position` | `"right"`, `"bottom"`, `"top"` or `"none"` | right |
| `bar_width` | Bar width; 1 leaves no gap | 0.85 |
| `outline_colour`, `outline_width` | Segment outlines; `NA` for none | grey20, 0.1 |
| `x_text_angle` | Sample label angle | 90 |
| `base_size` | Base font size (pt) | 9 |
| `y_label` | Y axis title | from the value type |

A legend below the bars suits many samples with long taxon names:

```{r bottom}
barchart.gg <- gmaio::plot_stacked_barchart(genus.p, metadata.df, project.l$palettes$taxa,
                                            top_n = 12, top_method = "mean",
                                            facet_variable = "Treatment",
                                            annotation_variables = "Time",
                                            annotation_colours = project.l$palettes,
                                            legend_title = "Genus", legend_position = "bottom", legend_ncol = 3,
                                            bar_width = 1, outline_colour = NA)
```

```{r bottom-show, echo = FALSE}
show_plot(barchart.gg)
```

The figure is sized for its samples and legend, so changing these needs no change to `save_plot()`.
Without annotation strips the result is an ordinary ggplot, so other ggplot2 changes can be added, e.g. `barchart.gg + ggplot2::labs(title = "sylph genera")`.

## Other profiles

Any profile works, as long as it is at the level you want to show.
SingleM at phylum level:

```{r singlem}
singlem.p <- gmaio::get_profile(project.l, "singlem_relative") |>
  gmaio::aggregate_profile(rank = "phylum")
barchart.gg <- gmaio::plot_stacked_barchart(singlem.p, metadata.df, project.l$palettes$taxa,
                                            top_n = 6, facet_variable = "Treatment", legend_title = "Phylum")
```

```{r singlem-show, echo = FALSE}
show_plot(barchart.gg)
```

For function profiles, aggregate by an annotation column (`aggregate_profile(profile, by = "Category")`) and convert to percentages with `as_relative_abundance()`, as `code/functional_profiles.R` does for CAZyme substrates.

## Colours

### Where colours come from

`main.R` assigns colours once and saves them to `palettes.yml` in the project directory, so a taxon or group keeps its colour in every figure and every run, including runs after more samples are added.
For every variable and for taxa, colours are taken in this order:

1. `colours:` in `config.yml`.
2. `palettes.yml`, which you can edit; your edits are kept on the next run.
3. New colours for anything not yet coloured, chosen to differ from those already in use.

`project.l$palettes` holds them all: one named vector per metadata variable and one for taxa (by label, at every rank).

```{r palettes}
names(project.l$palettes)
project.l$palettes$Treatment
utils::head(project.l$palettes$taxa, 4)
gmaio::special_colours()
```

`Other`, `Unassigned`, missing values and reference genomes always get these fixed greys, so they never look like a real group.

### Taxa colour schemes

Phyla get the most distinct colours in order of abundance.
Below phylum there are two schemes, set by `outputs: taxa_colour_scheme` in `config.yml`:

- `distinct` (default): every taxon gets its own colour in order of overall abundance, so the most abundant taxa are easiest to tell apart.
- `hierarchical`: taxa get shades of their phylum colour, which groups them visually but gives similar shades when a phylum has many taxa.

`assign_taxa_colours()` makes either scheme; the hierarchical one for the genera above looks like this:

```{r hierarchical}
taxa.df <- data.frame(Label = genus.p$features$Label, Phylum = genus.p$features$Phylum,
                      Abundance = apply(genus.p$values, 1, max))
hierarchical.v <- gmaio::assign_taxa_colours(taxa.df, scheme = "hierarchical")
barchart.gg <- gmaio::plot_stacked_barchart(genus.p, metadata.df, hierarchical.v,
                                            top_n = 12, top_method = "mean",
                                            facet_variable = "Treatment", legend_title = "Genus")
```

```{r hierarchical-show, echo = FALSE}
show_plot(barchart.gg)
```

In a project, set the scheme in `config.yml` and delete the taxa entries of `palettes.yml` (or the whole file) before rerunning `main.R`, since saved colours are kept.

### Changing colours

For one figure, pass modified colours:

```{r override}
taxa.v <- project.l$palettes$taxa
taxa.v[["(p__Verrucomicrobiota) g__Akkermansia"]] <- "#D7191C"
barchart.gg <- gmaio::plot_stacked_barchart(genus.p, metadata.df, taxa.v, top_n = 12, top_method = "mean")
```

To change them everywhere, set them in `config.yml`, using the labels as they appear in figures:

```yaml
colours:
  Treatment: {Control: "#4E79A7", Treated: "#E15759"}
  taxa: {"(p__Verrucomicrobiota) g__Akkermansia": "#D7191C", p__Bacillota: "#59A14F"}
```

or edit `palettes.yml` directly, then rerun `main.R`.
`get_palette(project.l, "taxa", labels)` looks colours up and gives any taxon without one a new colour; plot functions do this for you.
To colour a new metadata column, add it to `metadata: colour_variables` in `config.yml`.

`qualitative_palette(n)` and `assign_colours(values)` give the generated colours on their own, for your own figures.

## Saving

```r
gmaio::save_plot(barchart.gg, gmaio::figure_path(project.l$config, "barcharts", "sylph_taxonomic__genus"))
```

The size is set from the number of samples and legend entries; pass `width` and `height` (cm) to fix it instead.
