---
title: "Comparison samples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison samples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
# Show each figure exactly as save_plot() writes it, at its automatic size
show_plot <- function(plot, ...){
  path.s <- knitr::fig_path(".png")
  gmaio::save_plot(plot, path.s, dpi = 110, ...)
  knitr::include_graphics(path.s, dpi = 110)
}
```

The pipeline can profile reads from another study next to your own (`--comparison_reads`, and `--comparison_assemblies` to add their genes to the gene catalogue).
These comparison samples are not assembled or binned: they get sylph and SingleM profiles, are mapped to your dereplicated bins, and are mapped to the gene catalogue.
`code/comparison.R` puts them side by side with your samples, to show how different your samples are from each other and from the comparison samples.
This vignette follows that script on the synthetic example project, where 16 mice from two other facilities are the comparison samples.

```{r project}
project.l <- gmaio::example_project()
```

## Configuration

Comparison samples come from other studies, so they have their own metadata, set in the `comparison:` section of `config.yml`:

```yaml
comparison:
  metadata:
    file: data/comparison_metadata.xlsx
    sheet: 1                       # a sheet name or number
    sample_id_column: Sample_ID    # must match the --comparison_reads samplesheet `sample` column
    label_column: Name
    exclude_samples: []
  group_variable: Facility         # comparison metadata column to group comparison samples by
  dataset_labels: {study: This study, comparison: Other facilities}
```

The metadata can list more samples than were sequenced; only those in the pipeline outputs are used, and a comparison sample missing from it is an error.
`main.R` loads them with `add_comparison()`, after the MAGs and gene catalogue functions.

## Combined metadata and profiles

`combined_metadata()` stacks the two sets of samples.
`Dataset` says where each sample comes from, and `Comparison_group` holds the study's primary group variable for study samples and `comparison: group_variable` for comparison samples, with the study's levels first.
A comparison group named like a study group (for example the same site) is suffixed with the comparison dataset label, so the two datasets' groups are never pooled.

```{r metadata}
metadata.df <- gmaio::combined_metadata(project.l)
table(metadata.df$Dataset, metadata.df$Comparison_group)
```

`combined_profile()` joins a comparison profile with the study profile of the same name.
Taxa found in only one dataset are zero in the other.
Function profiles (`functions_ko`, ...) are joined with the study profiles of the catalogue the comparison reads were mapped to, the expanded one when the pipeline built it.

```{r profiles}
gmaio::combined_profile_names(project.l)
genus.p <- gmaio::aggregate_profile(gmaio::combined_profile(project.l, "sylph_taxonomic"), rank = "genus")
genus.p
```

## Ordination and PERMANOVA

The script ordinates each combined dataset, colouring samples by `Comparison_group` and shaping them by `Dataset`, and tests both variables with PERMANOVA.
Pairwise PERMANOVA between all groups shows which study groups differ from which comparison groups.

```{r ordination}
filtered.p <- gmaio::filter_profile(genus.p, min_prevalence = 3)
ordination <- gmaio::run_ordination(filtered.p, "pca_rclr")
permanova.df <- gmaio::run_permanova(ordination$distance, metadata.df, "Comparison_group", permutations = 999)
permdisp.l <- gmaio::run_permdisp(ordination$distance, metadata.df, "Comparison_group", permutations = 999)
ordination.gg <- gmaio::plot_ordination(ordination, metadata.df, colour_by = "Comparison_group",
                                        colours.v = project.l$palettes$Comparison_group, shape_by = "Dataset",
                                        ellipse = TRUE, title = "sylph genus",
                                        caption = gmaio::permanova_caption(permanova.df, permdisp.l))
```

```{r ordination-show, echo = FALSE}
show_plot(ordination.gg, width = 18, height = 13)
```

```{r pairwise}
gmaio::run_pairwise_permanova(ordination$distance, metadata.df, "Comparison_group", permutations = 999)
```

## Distances to each group

`comparison_distances()` measures, for every study sample, how far it is from each group separately.
It takes the sample's distance (in the ordination's distance, here robust Aitchison) to every sample of one group and averages them, then does the same for the next group, and so on.
For the sample's own group, the sample itself is left out.
So each study sample gives one value per group: its mean distance to the other samples of its own Location, and its mean distance to the samples of each other Location and of each comparison group.

The figure has one panel per study group (`From <group>`) and one box per group the distances go to (`To samples of`); each point is one study sample.
A low box for the sample's own group and higher boxes for the rest means the study groups are distinct; a comparison group with a box as low as the own-group box holds samples as similar as the study group's own.
Points from one sample are not independent, so the figure draws no tests; the PERMANOVA results are the inference.

```{r distances}
distances.df <- gmaio::comparison_distances(ordination$distance, metadata.df)
head(distances.df)
distances.gg <- gmaio::plot_comparison_distances(distances.df, project.l$palettes$Comparison_group,
                                                 y_label = paste("Mean distance,", ordination$distance_label))
```

```{r distances-show, echo = FALSE}
show_plot(distances.gg)
```

## Composition

Barcharts are split by dataset, with a strip of groups under the bars.
With many samples, `top_method = "mean"` keeps the legend to the taxa with the highest mean instead of each sample's own top taxa.

```{r barchart}
barchart.gg <- gmaio::plot_stacked_barchart(genus.p, metadata.df, project.l$palettes$taxa, top_n = 12, top_method = "mean",
                                            facet_variable = "Dataset", annotation_variables = "Comparison_group",
                                            annotation_colours = list(Comparison_group = project.l$palettes$Comparison_group),
                                            legend_title = "Genus")
```

```{r barchart-show, echo = FALSE}
show_plot(barchart.gg)
```

## Alpha diversity

Statistics on the figure can be left off with `show_test = FALSE` and `brackets = FALSE` (or `show_statistics.b <- FALSE` at the top of `code/comparison.R`); the tests are still returned.

```{r diversity}
diversity.l <- gmaio::plot_alpha_diversity(gmaio::alpha_diversity(genus.p), metadata.df, "Comparison_group",
                                           project.l$palettes$Comparison_group)
```

```{r diversity-show, echo = FALSE}
show_plot(diversity.l$plot)
```

Richness depends on sequencing depth.
Comparison samples are often sequenced to a different depth than the study's, so compare richness between datasets only when depths are similar, or use the depth-adjusted datasets.
`code/comparison.R` computes alpha diversity for the datasets of `code/diversity.R` that the comparison samples have, including SingleM marker gene reads rarefied, SingleM marker gene sequences (OTUs) rarefied and sylph approximated at a common depth, plus the MAG read counts rarefied (`mags_rarefied`), and plots each measure against depth, coloured by dataset.
It writes them to `Study_vs_comparison_alpha_diversity.xlsx` and the `alpha_diversity__<dataset>` and `alpha_diversity__<dataset>__vs_depth` figures.
`prokaryotic_depth(project.l, include_comparison = TRUE)` gives the depth of both datasets; `vignette("diversity", package = "gmaio")` explains the depth-adjusted measures and `exclude_unresolved`.
The comparison reads were mapped to the study's dereplicated bins only, so MAG measures count only the genomes the study recovered and are lower in the comparison samples by design.

## Detecting the study's MAGs

The pipeline maps comparison reads with CoverM to the study's dereplicated bins: the representatives of all bins of any quality after dereplication at 95% ANI (`<n>_dereplicated_bins`, the `derep_bins` set), the same genomes the study samples are mapped to in `<n>_coverm_bins`.
Reference genomes (`--reference_genomes`) are not part of this mapping, whatever the settings; they reach the comparison samples through the gene catalogue, when the run has `--reference_genomes_in_catalogue`, and through strain comparison, when the strain set includes them and the run has `--strain_include_comparison_reads` (see the strains vignette).
`mag_detection()` counts, for each genome, the samples of each group in which reads cover at least `min_covered_fraction` of it; requiring breadth guards against reads from relatives that hit only conserved regions.
The table goes to `Study_vs_comparison_MAG_detection.xlsx`, and the script draws the covered fraction of the HQ bins as a heatmap.

```r
detection.df <- gmaio::mag_detection(project.l, min_covered_fraction = 0.3)
covered.p <- gmaio::combined_profile(project.l, "mags_derep_bins_covered_fraction")
```

The example project has no CoverM mapping, so this part is not run here.
