---
title: "Differential abundance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Differential abundance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# The methods are optional packages; without them the code is shown but not run
methods_installed.b <- all(vapply(c("maaslin3", "MicrobiomeStat", "mixOmics"), requireNamespace, logical(1), quietly = TRUE))
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center", eval = methods_installed.b)
# Show each figure exactly as save_plot() writes it, at its automatic size
show_plot <- function(plot, ...){
  path.s <- knitr::fig_path(".png")
  gmaio::save_plot(plot, path.s, dpi = 110, ...)
  knitr::include_graphics(path.s, dpi = 110)
}
```

Differential abundance asks which features (taxa, MAGs or functions) differ between groups.
No single method is reliable for compositional, sparse microbiome data, so `code/differential_abundance.R` runs three different ones and keeps the features they agree on:

| Method | Approach | Effect reported |
|---|---|---|
| MaAsLin3 | Linear models on log abundance (`abundance` model) and on presence (`prevalence` model) | log2 coefficient |
| LinDA | Linear models on centred log-ratios with a correction for compositional bias | log2 fold change |
| sPLS-DA (mixOmics) | Sparse discriminant analysis on centred log-ratios, features picked by cross-validation | loading and selection stability |

The three are in `Suggests`; install them with `BiocManager::install(c("maaslin3", "MicrobiomeStat", "mixOmics"))`.
This vignette follows the script, then goes through the consensus and the figures.
It uses the synthetic example project; see `vignette("gmaio", package = "gmaio")` for how projects and profiles work.

```{r project}
project.l <- gmaio::example_project()
metadata.df <- gmaio::analysis_metadata(project.l)
```

## Choosing the data

The script tests genus-level sylph and SingleM profiles, the HQ MAGs and the KEGG and CAZy function profiles.
Rare features are removed first: those never reaching 0.01% and those present in fewer than 10% of samples.
They carry little information and add to the multiple-testing burden.

```{r data}
genus.p <- gmaio::get_profile(project.l, "sylph_taxonomic") |>
  gmaio::aggregate_profile(rank = "genus") |>
  gmaio::filter_profile(min_abundance = 0.01) |>
  gmaio::filter_profile(min_prevalence = 0.1)
genus.p
```

## The design

In the example, the treatment started after the week 0 samples, so treated and control mice differ only at week 4.
Testing Treatment across all samples would dilute that effect with the week 0 samples, so here the test uses week 4 only:

```{r week4}
week4.df <- metadata.df[metadata.df$Time == "Week_4", ]
table(week4.df$Treatment)
```

Think about the design before running the script: which samples answer the question, what else varies between the groups (`covariates`), and whether samples are repeated measures of the same subject (`random_effects`, see below).

## Running the methods

```{r run, message = FALSE}
da.l <- gmaio::run_differential_abundance(genus.p, week4.df, "Treatment",
                                          methods = c("maaslin3", "linda", "splsda"),
                                          min_prevalence = 0.1, splsda_repeats = 20, seed = 1234)
```

`da.l$results` has every feature from every method in one table, significant or not:

```{r results}
results.df <- da.l$results
head(results.df[results.df$Method == "LinDA", c("Method", "Label", "Contrast", "Reference", "Effect", "Q_value", "Enriched_in")])
table(results.df$Method, results.df$Model)
```

| Column | Meaning |
|---|---|
| `Method`, `Model` | Method, and for MaAsLin3 the abundance or prevalence model (sPLS-DA: the component) |
| `Contrast`, `Reference` | The level compared, and the reference level it is compared with |
| `Effect`, `Effect_type`, `SE` | Effect size and its kind (see the table above) |
| `P_value`, `Q_value` | Raw and FDR-adjusted p values (MaAsLin3 and LinDA) |
| `Stability` | How often sPLS-DA picked the feature across cross-validation runs |
| `Enriched_in` | The group in which the feature is higher |

The reference level is the first factor level, set by `reference_levels` in `config.yml`.
With more than two groups, MaAsLin3 and LinDA report one contrast per non-reference level.

`splsda_repeats` sets the cross-validation repeats for sPLS-DA; the script uses 50, which is slower but gives more stable selection.

## Consensus

`da_consensus()` keeps the features each method calls significant: MaAsLin3 (abundance model) and LinDA with `Q_value <= alpha` (0.05), and sPLS-DA with `Stability >= min_stability` (0.7).
Methods agree when they call a feature higher in the same group:

```{r consensus}
consensus.df <- gmaio::da_consensus(da.l$results, alpha = 0.05, min_stability = 0.7)
consensus.df[, c("Label", "Enriched_in", "MaAsLin3_q", "LinDA_q", "sPLSDA_stability", "N_methods")]
```

`N_methods` counts the methods agreeing, and `Conflicting_direction` flags features called higher in different groups by different methods.
Features found by two or three methods are the ones to report; those found by one are worth a look but less certain.
The five genera that increase with treatment in the synthetic data are all found by at least two methods.
Of the five that decrease, only two are found, each by one method: treated mice lose them entirely (zeros), which the log-ratio methods handle differently and which leaves little to test with six mice per group.

`code/differential_abundance.R` writes every result to `Differential_abundance_all_results.xlsx` and the consensus to `Differential_abundance_consensus.xlsx`.

## Figures

### Effects

`plot_da_consensus()` shows the effect from each method for features found by at least `min_methods`:

```{r effects}
effects.gg <- gmaio::plot_da_consensus(consensus.df, project.l$palettes$Treatment, min_methods = 1)
```

```{r effects-show, echo = FALSE}
show_plot(effects.gg)
```

The effect scales differ between methods, so each panel has its own x axis.
Options: `methods` (panels shown, in order), `max_features`, `bar_width`, `label_width` (wrap long labels), `x_label`, `legend_title`, `base_size` and `legend_position`.

### Heatmap

`plot_da_heatmap()` shows the abundance of every feature called by at least `min_methods` (default 1), with tracks for the group each method called it for, so disagreement between methods shows as a colour mismatch:

```{r heatmap}
heatmap.ht <- gmaio::plot_da_heatmap(genus.p, consensus.df, week4.df, "Treatment", project.l$palettes)
```

```{r heatmap-show, echo = FALSE}
show_plot(heatmap.ht)
```

Rows are split by the group they are higher in and columns by the tested group.
It is built on `plot_heatmap()`, so all of its options apply (see `vignette("heatmaps", package = "gmaio")`), for example:

```{r heatmap-options}
heatmap.ht <- gmaio::plot_da_heatmap(genus.p, consensus.df, week4.df, "Treatment", project.l$palettes,
                                     min_methods = 2, row_labels = "Genus", colours = "Rocket",
                                     cell_size = 0.5, row_split = NULL, column_annotations = "Cage")
```

```{r heatmap-options-show, echo = FALSE}
show_plot(heatmap.ht)
```

`not_called_colour` and `conflict_colour` set the track colours, and `rank_annotation` the taxonomy track (`"Phylum"` by default).

### Boxplots

`plot_feature_boxplots()` shows the samples behind each call.
The script draws the features found by at least two methods, 12 per page:

```{r boxplots}
agreed.v <- unique(consensus.df$Feature_ID[consensus.df$N_methods >= 2])
boxplots.gg <- gmaio::plot_feature_boxplots(genus.p, metadata.df, agreed.v, "Treatment", project.l$palettes$Treatment,
                                            shape_by = "Time", ncol = 3)
```

```{r boxplots-show, echo = FALSE}
show_plot(boxplots.gg)
```

All samples are shown here, with week 0 and week 4 as shapes, which makes it clear the difference arises at week 4.
Zeros are drawn at half the smallest non-zero value on the log scale; `log_scale = FALSE` shows the values as they are.
Other options: `shapes`, `box_width`, `box_alpha`, `point_size`, `jitter_width`, `label_width` (panel title wrapping), `free_y`, `x_label`, `y_label`, `x_text_angle`, `base_size` and `legend_position`.

## Covariates and repeated measures

`covariates` adds fixed effects, such as age, sex or batch, that differ between groups and could explain differences.
`random_effects` handles repeated measures: here each mouse was sampled twice, so all samples can be used with a random intercept per mouse.
MaAsLin3 and LinDA fit mixed models; sPLS-DA cannot and uses the tested variable only.

```{r mixed, message = FALSE}
mixed.l <- gmaio::run_differential_abundance(genus.p, metadata.df, "Treatment", covariates = "Time",
                                             random_effects = "Mouse", methods = c("maaslin3", "linda"))
gmaio::da_consensus(mixed.l$results)[, c("Label", "Enriched_in", "N_methods")]
```

With Treatment and Time as separate terms, the treatment effect is averaged over both time points, so fewer genera pass than in the week 4 test above.
The model that matches the example design would include a Treatment by Time interaction, which needs `maaslin3::maaslin3()` or `MicrobiomeStat::linda()` called directly (the profile's `values` and the metadata are their inputs).

In a project, covariates come from `analysis: covariates` in `config.yml`.
For random effects, add `random_effects = "Mouse"` to the `gmaio::run_differential_abundance()` call in `code/differential_abundance.R`.

## Running one method

Each method can be run on its own with the same arguments, returning the same columns: `run_maaslin3()`, `run_linda()` and `run_splsda()` (which also returns the fitted model and its cross-validated performance).
`run_linda()` leaves out features nonzero in fewer than `min_nonzero` (3) samples, which LinDA cannot test.

## Saving

```r
figure_path.f <- function(suffix.s) gmaio::figure_path(project.l$config, "differential_abundance", paste0("sylph_genus__", suffix.s))
gmaio::save_plot(effects.gg, figure_path.f("consensus"))
gmaio::save_plot(heatmap.ht, figure_path.f("heatmap"))
gmaio::save_plot(boxplots.gg, figure_path.f("boxplots"))
```

All three are sized automatically; pass `width` and `height` (cm) to fix the size.
