---
title: "Alpha diversity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Alpha diversity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
# Show each figure exactly as save_plot() writes it, at its automatic size
show_plot <- function(plot, ...){
  path.s <- knitr::fig_path(".png")
  gmaio::save_plot(plot, path.s, dpi = 110, ...)
  knitr::include_graphics(path.s, dpi = 110)
}
```

Alpha diversity summarises each sample on its own: how many taxa it has and how evenly they are spread.
`code/diversity.R` computes it with `alpha_diversity()` and compares groups with `plot_alpha_diversity()`, which draws the figure and returns the test tables.
This vignette follows that script, explains the tests, and goes through the options for changing the figure.
It uses the synthetic example project; see `vignette("gmaio", package = "gmaio")` for how projects and profiles work.

```{r project}
project.l <- gmaio::example_project()
metadata.df <- gmaio::analysis_metadata(project.l)
```

## Measures

`alpha_diversity()` takes a profile and returns one row per sample:

```{r measures}
genus.p <- gmaio::get_profile(project.l, "sylph_taxonomic") |>
  gmaio::aggregate_profile(rank = "genus")
diversity.df <- gmaio::alpha_diversity(genus.p)
head(diversity.df)
```

| Measure | What it is |
|---|---|
| `Richness` | Number of taxa above zero |
| `Shannon` | Shannon entropy (natural log): richness and evenness together |
| `Simpson` | Gini-Simpson, 1 - sum of squared proportions: the chance two random reads come from different taxa |
| `Pielou` | Evenness, Shannon / log(Richness), from 0 (one taxon dominates) to 1 (all equal); only with `evenness = TRUE`, since it inherits the depth dependence of richness |
| `Chao1` | Estimated richness including unseen taxa; integer read counts only |

Shannon and Simpson are computed on proportions, so they suit relative abundance, coverage and counts alike.
Richness depends on sequencing depth: for read counts (e.g. the `mags_*_read_count` profiles), `rarefy_depth` subsamples every sample to the same number of reads first, dropping samples below it:

```r
counts.p <- gmaio::get_profile(project.l, "mags_hq_derep_bins_read_count")
diversity.df <- gmaio::alpha_diversity(counts.p, rarefy_depth = 100000, seed = 1234)
```

Relative abundance profiles from sylph or SingleM cannot be rarefied, since they have no read counts; see [Sequencing depth](#sequencing-depth) for the ways round that.

## What a feature is in each profile

Every measure counts or weighs features, so what a feature is decides what the diversity means:

| Profile | A feature is | Values |
|---|---|---|
| `sylph_taxonomic` | A genome in sylph's database (GTDB) whose k-mers the sample's reads contain closely enough; summed to a rank with `aggregate_profile()` | Taxonomic abundance: the estimated share of cells |
| `singlem_relative` | A lineage from SingleM's single-copy marker genes, at the deepest rank SingleM could assign | Marker gene coverage, as percent |
| `singlem_read_count` | The same lineages, from the reads of each marker gene sequence (OTU) | Marker gene reads |
| `mags_<set>_relative_abundance` | A bin of the study | CoverM relative abundance of the reads mapped to the set |
| `mags_<set>_read_count` | A bin of the study | Reads mapped (CoverM) |

sylph only sees species in its database, so a species absent from GTDB (common in less studied hosts) is missing from sylph however abundant it is.
SingleM classifies the marker gene reads themselves, so it sees such species, but often only to family or genus.
MAG profiles count only the genomes the study recovered.

### Unresolved lineages

When a profile is aggregated to a rank, reads or genomes that could not be classified to that rank are not dropped: they are summed under their deepest resolved parent, and that sum is one feature (`Unclassified (p__Bacillota_A) f__Lachnospiraceae` at genus level).
Each therefore counts as one taxon in richness and enters Shannon and Simpson as one, however many taxa it stands for.
This is minor for sylph, whose features are database genomes with a full taxonomy, but SingleM reads often stop above genus or species.
`unresolved_features()` shows which features they are, and `exclude_unresolved = TRUE` leaves them out, so the measures describe the taxa resolved at the rank:

```{r unresolved}
singlem_genus.p <- gmaio::get_profile(project.l, "singlem_read_count") |>
  gmaio::aggregate_profile(rank = "genus")
table(gmaio::unresolved_features(singlem_genus.p))
resolved.df <- gmaio::alpha_diversity(singlem_genus.p, exclude_unresolved = TRUE)
summary(resolved.df$Richness)
```

`code/diversity.R` and `code/comparison.R` have `exclude_unresolved.b` at the top (`FALSE` by default).
With rarefied counts, samples are rarefied first, with all their reads, and the unresolved features are left out after.

## Sequencing depth

Deeper sequencing detects more rare taxa, so richness rises with depth, and Shannon somewhat with it; Simpson, driven by the dominant taxa, hardly moves.
When groups or datasets were sequenced to different depths, part of a richness difference can be depth.

### Depth per sample

`prokaryotic_depth()` gives SingleM's estimate of the bacterial and archaeal bases in each sample, after read QC and host removal.
Profilers can only detect a prokaryote from prokaryotic reads, so this is the depth that limits detection: a sample with many host or plant reads has less of it than its read count suggests.

```{r depth}
depth.v <- gmaio::prokaryotic_depth(project.l)
summary(depth.v)
```

### How much a measure follows depth

`plot_diversity_vs_depth()` plots each measure against depth, with the Spearman correlation per panel; `code/diversity.R` writes one such figure per dataset (`<dataset>__vs_depth`) and the correlations to the `Depth_correlation` sheet.

```{r depth-plot}
singlem_genus.df <- gmaio::alpha_diversity(singlem_genus.p)
depth.l <- gmaio::plot_diversity_vs_depth(singlem_genus.df, depth.v, metadata.df, "Treatment",
                                          project.l$palettes$Treatment)
depth.l$correlation
```

```{r depth-plot-show, echo = FALSE}
show_plot(depth.l$plot)
```

### SingleM: rarefied marker gene reads

The pipeline's SingleM OTU table counts the reads on each marker gene sequence in each sample.
gmaio sums them by lineage into `singlem_read_count`, which `alpha_diversity()` rarefies like any read count profile.
Subsampling the pooled marker gene reads is equivalent to subsampling the sequencing reads, so the rarefied measures compare samples at the same depth.
Unlike the condensed profile, which SingleM builds by combining the marker genes, each read keeps the taxonomy of its own marker gene sequence, so the two profiles differ a little.

```{r singlem-rarefied}
rarefied.df <- gmaio::alpha_diversity(singlem_genus.p, rarefy_depth = 700,
                                      exclude_unresolved = TRUE)
head(rarefied.df)
```

`singlem_otu_diversity()` measures the marker gene sequences themselves (OTUs), rarefied the same way, with no reference database: taxa missing from GTDB count too.
Each organism has one sequence per marker gene, so richness is given per marker gene (averaged over them), which approximates the number of distinct populations.

```{r singlem-otus}
otus.df <- gmaio::singlem_otu_diversity(project.l$tables$singlem_otus, rarefy_depth = 700)
head(otus.df)
```

### sylph: an approximation at a common depth

sylph reports abundances and coverages, not reads, so its profiles cannot be rarefied.
Profiling reads subsampled to the same number is the exact answer, and needs the pipeline to run sylph again.
`sylph_profile_at_depth()` approximates it from the pipeline's `sylph_profile.tsv`: each genome's k-mer coverage is scaled by the sample's depth ratio, its matched k-mers by the share still expected at that coverage (1 - exp(-coverage)), and genomes left with fewer than sylph's minimum of 50 matched k-mers are dropped.
It ignores sylph's ANI threshold, which can drop a few more genomes at low coverage, so it slightly overestimates richness at the target depth: use it as a sensitivity check.

```{r sylph-depth}
sylph_depth.p <- gmaio::sylph_profile_at_depth(project.l$tables$sylph_profile, depth.v,
                                               project.l$profiles$sylph_taxonomic$features)
attr(sylph_depth.p, "target_depth")
sylph_depth.df <- gmaio::alpha_diversity(gmaio::aggregate_profile(sylph_depth.p, rank = "species"))
summary(sylph_depth.df$Richness)
```

### In the template scripts

`code/diversity.R` runs these next to the plain profiles: `singlem_rarefied__genus` and `__species` (SingleM marker gene reads rarefied), `singlem_otus_rarefied` and `sylph_at_depth__species`.
Samples are rarefied to the smallest sample with at least `min_singlem_reads.n` marker gene reads, and sylph is scaled to the smallest prokaryotic depth of at least `min_prokaryotic_bases.n`; shallower samples are left out of those datasets.
`code/comparison.R` does the same for the study and comparison samples together, and also rarefies the MAG read counts (`mags_rarefied`).
Shannon and Simpson of the plain profiles, and the depth-adjusted datasets, are the ones to compare between datasets sequenced to different depths.

## The template figure

`code/diversity.R` does this for sylph and SingleM species, SingleM genera, the HQ MAGs and the depth-adjusted datasets above, comparing the primary group:

```{r template}
diversity.l <- gmaio::plot_alpha_diversity(diversity.df, metadata.df, "Treatment", project.l$palettes$Treatment)
```

```{r template-show, echo = FALSE}
show_plot(diversity.l$plot)
```

Each panel is one measure with its own y axis; boxes show the median and quartiles and points the samples.
Groups are drawn in factor level order, so `reference_levels` in `config.yml` puts the reference group first.
In a project, edit the `gmaio::plot_alpha_diversity()` call in `code/diversity.R` and rerun the script.

## Tests

With two groups every panel shows a Wilcoxon rank-sum test; with more it shows a Kruskal-Wallis test.
`diversity.l$tests` has them all, with a BH adjustment across the measures in `P_adjusted`:

```{r tests}
diversity.l$tests
```

In the example, treated mice lose rare genera by week 4, but pooling both time points hides much of that.
Comparing the four Treatment and Time combinations shows it, and with more than two groups the pairs that differ get brackets:

```{r groups}
metadata.df$Group <- factor(paste(metadata.df$Treatment, metadata.df$Time))
diversity.l <- gmaio::plot_alpha_diversity(diversity.df, metadata.df, "Group")
```

```{r groups-show, echo = FALSE}
show_plot(diversity.l$plot)
```

The pairwise tests are Dunn's test by default, the standard follow-up to Kruskal-Wallis: it ranks all samples together, so it agrees with the omnibus test.
P values are BH adjusted within each measure, and only pairs with an adjusted p below `bracket_alpha` (0.05) get a bracket, marked `*` (< 0.05), `**` (< 0.01) or `***` (< 0.001).
`diversity.l$pairwise` has every pair, with group sizes, means and medians:

```{r pairwise}
pairwise.df <- diversity.l$pairwise
pairwise.df[pairwise.df$Measure == "Richness", c("Group_1", "Group_2", "N_1", "N_2", "Mean_1", "Mean_2",
                                                 "Statistic", "P_adjusted", "Significance")]
```

`code/diversity.R` writes the values, tests and pairwise tables to `Alpha_diversity.xlsx`.

The tests treat samples as independent.
Here each mouse was sampled twice, so for a strict comparison test one time point at a time (`metadata.df[metadata.df$Time == "Week_4", ]`), or model the repeated measures yourself from `diversity.df`.

## Test options

| Argument | Controls | Default |
|---|---|---|
| `pairwise_test` | `"dunn"` or `"wilcoxon"` (separate rank-sum test per pair) | dunn |
| `p_adjust_method` | Correction within each measure, see `stats::p.adjust()` | BH |
| `brackets` | Draw brackets for significant pairs | `TRUE` |
| `bracket_alpha` | Adjusted p below which a pair gets a bracket | 0.05 |
| `bracket_label` | `"stars"` or `"p"` (the adjusted p value) | stars |
| `bracket_size` | Bracket label size (pt) | 8 |
| `show_test` | Print the panel test above the boxes | `TRUE` |

```{r test-options}
diversity.l <- gmaio::plot_alpha_diversity(diversity.df, metadata.df, "Group",
                                           measures = c("Richness", "Shannon"),
                                           pairwise_test = "wilcoxon", bracket_label = "p")
```

```{r test-options-show, echo = FALSE}
show_plot(diversity.l$plot)
```

### Leaving the statistics off the figure

`show_test = FALSE` drops the panel test (Wilcoxon or Kruskal-Wallis p value) and `brackets = FALSE` the pairwise brackets.
The tests are still run and returned (`diversity.l$tests`, `diversity.l$pairwise`), so the tables keep them.
Every template script that draws boxplots by group has a `show_statistics.b` switch at the top that sets both: `diversity.R`, `comparison.R`, `strains.R`, `genomespot.R`, `nonpareil.R` and `mobile_elements.R`.
The same two arguments work for every boxplot function built on `plot_group_boxplots()`, such as `plot_nonpareil_metrics()`, `plot_community_traits()` and `plot_instrain_diversity()`.

```{r no-statistics}
diversity.l <- gmaio::plot_alpha_diversity(diversity.df, metadata.df, "Group", measures = c("Richness", "Shannon"),
                                           show_test = FALSE, brackets = FALSE)
```

```{r no-statistics-show, echo = FALSE}
show_plot(diversity.l$plot)
```

## Figure options

| Argument | Controls | Default |
|---|---|---|
| `measures` | Measures shown, in order | Richness, Shannon, Simpson |
| `panel_labels` | Panel titles, e.g. `c(Richness = "Observed genera")` | measure names |
| `shape_by`, `shapes` | Point shape by a metadata column, and the shapes | none |
| `box_width`, `box_alpha` | Box width and fill transparency | 0.6, 0.5 |
| `point_size`, `point_alpha`, `jitter_width` | Points | 1.8, 1, 0.12 |
| `ncol` | Panel columns | up to 4 |
| `free_y` | A y axis per panel | `TRUE` |
| `x_label`, `y_label` | Axis titles | none |
| `x_text_angle` | Group label angle | 30 |
| `base_size`, `legend_position` | Font size (pt) and legend position | 9, right |

For example, Treatment on the x axis with time point as the point shape, two panels per row and clearer panel titles:

```{r figure-options}
diversity.l <- gmaio::plot_alpha_diversity(diversity.df, metadata.df, "Treatment", project.l$palettes$Treatment,
                                           shape_by = "Time", ncol = 2, x_text_angle = 0,
                                           panel_labels = c(Richness = "Observed genera", Shannon = "Shannon index",
                                                            Simpson = "Gini-Simpson index"))
```

```{r figure-options-show, echo = FALSE}
show_plot(diversity.l$plot)
```

The result is a ggplot, so further ggplot2 changes can be added to `diversity.l$plot`.

## Other per-sample values

`plot_group_boxplots()` draws the same figure, with the same tests, for any per-sample values in long format (a `Value` column and one column naming the panel).
For example, the share of reads SingleM assigns to bacteria and archaea:

```{r other}
fraction.df <- merge(project.l$tables$singlem_fraction[, c("Sample_ID", "read_fraction")], metadata.df, by = "Sample_ID")
fraction.df$Measure <- "Prokaryotic fraction (%)"
fraction.df$Value <- fraction.df$read_fraction
boxplots.l <- gmaio::plot_group_boxplots(fraction.df, "Measure", "Group")
```

```{r other-show, echo = FALSE}
show_plot(boxplots.l$plot)
```

`plot_nonpareil_metrics()` uses it for Nonpareil coverage and diversity, and takes the same options.

## Saving

```r
gmaio::save_plot(diversity.l$plot, gmaio::figure_path(project.l$config, "diversity", "sylph_taxonomic__genus"))
gmaio::write_xlsx_tables(list(Values = diversity.df, Tests = diversity.l$tests, Pairwise = diversity.l$pairwise),
                         gmaio::output_path(project.l$config, "tables", "Alpha_diversity.xlsx"))
```

The figure is sized from the number of panels and groups; pass `width` and `height` (cm) to fix it instead.
