---
title: "Alpha diversity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Alpha diversity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
# Show each figure exactly as save_plot() writes it, at its automatic size
show_plot <- function(plot, ...){
  path.s <- knitr::fig_path(".png")
  gmaio::save_plot(plot, path.s, dpi = 110, ...)
  knitr::include_graphics(path.s, dpi = 110)
}
```

Alpha diversity summarises each sample on its own: how many taxa it has and how evenly they are spread.
`code/diversity.R` computes it with `alpha_diversity()` and compares groups with `plot_alpha_diversity()`, which draws the figure and returns the test tables.
This vignette follows that script, explains the tests, and goes through the options for changing the figure.
It uses the synthetic example project; see `vignette("gmaio", package = "gmaio")` for how projects and profiles work.

```{r project}
project.l <- gmaio::example_project()
metadata.df <- gmaio::analysis_metadata(project.l)
```

## Measures

`alpha_diversity()` takes a profile and returns one row per sample:

```{r measures}
genus.p <- gmaio::get_profile(project.l, "sylph_taxonomic") |>
  gmaio::aggregate_profile(rank = "genus")
diversity.df <- gmaio::alpha_diversity(genus.p)
head(diversity.df)
```

| Measure | What it is |
|---|---|
| `Richness` | Number of taxa above zero |
| `Shannon` | Shannon entropy (natural log): richness and evenness together |
| `Simpson` | Gini-Simpson, 1 - sum of squared proportions: the chance two random reads come from different taxa |
| `Pielou` | Evenness, Shannon / log(Richness), from 0 (one taxon dominates) to 1 (all equal) |
| `Chao1` | Estimated richness including unseen taxa; integer read counts only |

Shannon, Simpson and Pielou are computed on proportions, so they suit relative abundance, coverage and counts alike.
Richness depends on sequencing depth: for read counts (e.g. the `mags_*_read_count` profiles), `rarefy_depth` subsamples every sample to the same number of reads first, dropping samples below it:

```r
counts.p <- gmaio::get_profile(project.l, "mags_hq_derep_bins_read_count")
diversity.df <- gmaio::alpha_diversity(counts.p, rarefy_depth = 100000, seed = 1234)
```

Relative abundance profiles from sylph or SingleM cannot be rarefied, since they have no read counts; their richness reflects each tool's detection threshold, which is applied the same way to every sample.

## The template figure

`code/diversity.R` does this for sylph and SingleM species, SingleM genera and the HQ MAGs, comparing the primary group:

```{r template}
diversity.l <- gmaio::plot_alpha_diversity(diversity.df, metadata.df, "Treatment", project.l$palettes$Treatment)
```

```{r template-show, echo = FALSE}
show_plot(diversity.l$plot)
```

Each panel is one measure with its own y axis; boxes show the median and quartiles and points the samples.
Groups are drawn in factor level order, so `reference_levels` in `config.yml` puts the reference group first.
In a project, edit the `gmaio::plot_alpha_diversity()` call in `code/diversity.R` and rerun the script.

## Tests

With two groups every panel shows a Wilcoxon rank-sum test; with more it shows a Kruskal-Wallis test.
`diversity.l$tests` has them all, with a BH adjustment across the measures in `P_adjusted`:

```{r tests}
diversity.l$tests
```

In the example, treated mice lose rare genera by week 4, but pooling both time points hides much of that.
Comparing the four Treatment and Time combinations shows it, and with more than two groups the pairs that differ get brackets:

```{r groups}
metadata.df$Group <- factor(paste(metadata.df$Treatment, metadata.df$Time))
diversity.l <- gmaio::plot_alpha_diversity(diversity.df, metadata.df, "Group")
```

```{r groups-show, echo = FALSE}
show_plot(diversity.l$plot)
```

The pairwise tests are Dunn's test by default, the standard follow-up to Kruskal-Wallis: it ranks all samples together, so it agrees with the omnibus test.
P values are BH adjusted within each measure, and only pairs with an adjusted p below `bracket_alpha` (0.05) get a bracket, marked `*` (< 0.05), `**` (< 0.01) or `***` (< 0.001).
`diversity.l$pairwise` has every pair, with group sizes, means and medians:

```{r pairwise}
pairwise.df <- diversity.l$pairwise
pairwise.df[pairwise.df$Measure == "Richness", c("Group_1", "Group_2", "N_1", "N_2", "Mean_1", "Mean_2",
                                                 "Statistic", "P_adjusted", "Significance")]
```

`code/diversity.R` writes the values, tests and pairwise tables to `Alpha_diversity.xlsx`.

The tests treat samples as independent.
Here each mouse was sampled twice, so for a strict comparison test one time point at a time (`metadata.df[metadata.df$Time == "Week_4", ]`), or model the repeated measures yourself from `diversity.df`.

## Test options

| Argument | Controls | Default |
|---|---|---|
| `pairwise_test` | `"dunn"` or `"wilcoxon"` (separate rank-sum test per pair) | dunn |
| `p_adjust_method` | Correction within each measure, see `stats::p.adjust()` | BH |
| `brackets` | Draw brackets for significant pairs | `TRUE` |
| `bracket_alpha` | Adjusted p below which a pair gets a bracket | 0.05 |
| `bracket_label` | `"stars"` or `"p"` (the adjusted p value) | stars |
| `bracket_size` | Bracket label size (pt) | 8 |
| `show_test` | Print the panel test above the boxes | `TRUE` |

```{r test-options}
diversity.l <- gmaio::plot_alpha_diversity(diversity.df, metadata.df, "Group",
                                           measures = c("Richness", "Pielou"),
                                           pairwise_test = "wilcoxon", bracket_label = "p")
```

```{r test-options-show, echo = FALSE}
show_plot(diversity.l$plot)
```

## Figure options

| Argument | Controls | Default |
|---|---|---|
| `measures` | Measures shown, in order | Richness, Shannon, Simpson, Pielou |
| `panel_labels` | Panel titles, e.g. `c(Richness = "Observed genera")` | measure names |
| `shape_by`, `shapes` | Point shape by a metadata column, and the shapes | none |
| `box_width`, `box_alpha` | Box width and fill transparency | 0.6, 0.5 |
| `point_size`, `point_alpha`, `jitter_width` | Points | 1.8, 1, 0.12 |
| `ncol` | Panel columns | up to 4 |
| `free_y` | A y axis per panel | `TRUE` |
| `x_label`, `y_label` | Axis titles | none |
| `x_text_angle` | Group label angle | 30 |
| `base_size`, `legend_position` | Font size (pt) and legend position | 9, right |

For example, Treatment on the x axis with time point as the point shape, two panels per row and clearer panel titles:

```{r figure-options}
diversity.l <- gmaio::plot_alpha_diversity(diversity.df, metadata.df, "Treatment", project.l$palettes$Treatment,
                                           shape_by = "Time", ncol = 2, x_text_angle = 0,
                                           panel_labels = c(Richness = "Observed genera", Shannon = "Shannon index",
                                                            Simpson = "Gini-Simpson index", Pielou = "Pielou evenness"))
```

```{r figure-options-show, echo = FALSE}
show_plot(diversity.l$plot)
```

The result is a ggplot, so further ggplot2 changes can be added to `diversity.l$plot`.

## Other per-sample values

`plot_group_boxplots()` draws the same figure, with the same tests, for any per-sample values in long format (a `Value` column and one column naming the panel).
For example, the share of reads SingleM assigns to bacteria and archaea:

```{r other}
fraction.df <- merge(project.l$tables$singlem_fraction[, c("Sample_ID", "read_fraction")], metadata.df, by = "Sample_ID")
fraction.df$Measure <- "Prokaryotic fraction (%)"
fraction.df$Value <- fraction.df$read_fraction
boxplots.l <- gmaio::plot_group_boxplots(fraction.df, "Measure", "Group")
```

```{r other-show, echo = FALSE}
show_plot(boxplots.l$plot)
```

`plot_nonpareil_metrics()` uses it for Nonpareil coverage and diversity, and takes the same options.

## Saving

```r
gmaio::save_plot(diversity.l$plot, gmaio::figure_path(project.l$config, "diversity", "sylph_taxonomic__genus"))
gmaio::write_xlsx_tables(list(Values = diversity.df, Tests = diversity.l$tests, Pairwise = diversity.l$pairwise),
                         gmaio::output_path(project.l$config, "tables", "Alpha_diversity.xlsx"))
```

The figure is sized from the number of panels and groups; pass `width` and `height` (cm) to fix it instead.
