---
title: "Getting started with gmaio"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with gmaio}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center", dpi = 96)
```

gmaio turns the output of the [genome_metagenome_all_in_one](https://github.com/julianzaugg/genome_metagenome_all_in_one) pipeline into tables, figures and statistics.
This guide explains how a project fits together and the handful of objects and functions everything else builds on.
The worked examples for individual figures are in the other vignettes:

- `vignette("heatmaps", package = "gmaio")`
- `vignette("ordination", package = "gmaio")`

## How a project fits together

A project is a directory with a `config.yml`, a metadata table and a `code/` folder of scripts, created by `gmaio::create_project()`.

1. `code/fetch_results.R` copies the pipeline outputs gmaio reads from the server (optional).
2. `code/main.R` does the bulk processing: it reads every available pipeline output, links samples to the metadata, assigns colours, writes the standard tables to `Result_tables/` and saves the processed project to `Result_other/processed.rds`.
3. Every other script in `code/` is one analysis.
   Each starts by loading the processed project, so the scripts are independent and can be rerun in any order.

The scripts contain no analysis logic of their own.
They choose which data to use and call `gmaio::` functions, so what a figure looks like is controlled by the arguments of those functions.
Every function has a help page (`?gmaio::plot_heatmap`) listing all of its arguments.
The scripts only set the few that differ from the defaults, so read the help page to see what else you can change.

## The processed project

An analysis script starts with:

```r
project.l <- gmaio::load_processed()
```

In these vignettes, `example_project()` builds the same object from a small synthetic dataset shipped with the package: 16 mouse samples, Control vs Treated, sampled at week 0 and week 4, profiled with sylph and SingleM.

```{r project}
project.l <- gmaio::example_project()
project.l
```

The project is a list:

| Element | What it holds |
|---|---|
| `config` | The parsed `config.yml` (paths, groups, ranks, permutations, colours) |
| `metadata` | The metadata table, linked to the pipeline sample names |
| `profiles` | Every abundance or function table, as `gm_profile` objects |
| `tables` | Other results (MAG summaries, SingleM prokaryotic fraction, ...) |
| `palettes` | Colours for metadata variables and taxa |

## Metadata

`analysis_metadata()` returns the metadata without excluded samples, in the order set by `sample_order` in `config.yml`.
gmaio adds `Sample_ID` (the ID used everywhere), `Sample_label` (the display label, from `label_column`) and `Excluded`.

```{r metadata}
metadata.df <- gmaio::analysis_metadata(project.l)
head(metadata.df)
```

The first of `analysis: group_variables` in the config is the primary group, used for splitting and colouring figures and for the statistical tests:

```{r group}
gmaio::primary_group(project.l)
```

## Profiles

Every abundance or function table is a `gm_profile`: a features x samples matrix (`values`) plus a `features` data frame with at least `Feature_ID` and a display `Label`.
Taxonomic profiles also have one column per rank (`Domain` to `Species`).
The same functions therefore work on sylph, SingleM, MAG and DRAM function profiles alike.

```{r profiles}
names(project.l$profiles)
profile <- gmaio::get_profile(project.l, "sylph_taxonomic")
profile
profile$values[1:3, 1:4]
profile$features[1:3, c("Feature_ID", "Phylum", "Genus", "Label")]
```

`get_profile()` drops excluded samples and, for relative abundance, rescales each sample back to 100%.

### Working with profiles

A figure usually starts with a few steps that choose the data, each taking and returning a profile.

| Function | What it does |
|---|---|
| `aggregate_profile(profile, rank = "genus")` | Sum features to a taxonomic rank, or `by =` any feature column |
| `filter_profile(profile, min_abundance, min_prevalence)` | Keep features reaching an abundance in any sample, or present in enough samples |
| `top_features(profile, n)` | Keep the `n` most abundant features (by mean or max) |
| `collapse_top_n(profile, top_n)` | Keep the top features and sum the rest into `"Other"` (for barcharts) |
| `subset_features()`, `subset_samples()` | Keep given features or samples |
| `as_relative_abundance()` | Rescale each sample to 100 |
| `profile_to_long()` | A long data frame with metadata joined, for your own ggplot figures |

For example, the ten most abundant genera present in at least four samples:

```{r verbs}
genus.p <- gmaio::get_profile(project.l, "sylph_taxonomic") |>
  gmaio::aggregate_profile(rank = "genus") |>
  gmaio::filter_profile(min_prevalence = 4) |>
  gmaio::top_features(10)
genus.p
round(rowMeans(genus.p$values), 1) |> stats::setNames(genus.p$features$Genus)
```

## Colours

Colours are assigned once, in `main.R`, and saved to `palettes.yml` in the project directory, so a group or taxon has the same colour in every figure and every run.
Colours are chosen in this order:

1. `colours:` in `config.yml`.
2. `palettes.yml`, which you can edit directly; your edits are kept.
3. Newly generated colours for anything not yet coloured.

```{r palettes}
project.l$palettes$Treatment
project.l$palettes$Time
gmaio::get_palette(project.l, "taxa", c("p__Bacteroidota", "p__Bacillota"))
```

Plot functions take either the whole palette list (`project.l$palettes`) or the colours for one variable (`project.l$palettes$Treatment`).

## Saving figures and tables

`save_plot()` saves ggplots and heatmaps with sizes in centimetres.
Heatmaps are sized automatically to fit their cells, labels and legends.
`figure_path()` builds a path in the project figure directory with the configured format (`outputs: figure_format`).
`write_xlsx_tables()` writes a named list of data frames to one formatted Excel file, one sheet each.

```r
gmaio::save_plot(figure.gg, gmaio::figure_path(project.l$config, "heatmaps", "sylph_genus"), width = 16, height = 12)
gmaio::write_xlsx_tables(list(PERMANOVA = permanova.df), gmaio::output_path(project.l$config, "tables", "Statistics.xlsx"))
```

## Scripts and their main functions

| Script | Main functions |
|---|---|
| `main.R` | `read_config()`, `start_project()`, `add_sylph()`, `add_singlem()`, `add_mags()`, `add_gene_catalogue_functions()`, `add_palettes()`, `write_project_tables()` |
| `barcharts.R` | `plot_stacked_barchart()` |
| `heatmaps.R` | `plot_heatmap()` |
| `ordination.R` | `run_ordination()`, `plot_ordination()`, `run_permanova()`, `run_permdisp()`, `plot_loadings()` |
| `diversity.R` | `alpha_diversity()`, `plot_alpha_diversity()` |
| `differential_abundance.R` | `run_differential_abundance()`, `da_consensus()`, `plot_da_consensus()` |
| `functional_profiles.R` | `plot_heatmap()`, `plot_stacked_barchart()` |
| `mag_summary.R` | `plot_mag_quality()`, `plot_mag_counts()`, `plot_mag_taxonomy()`, `within_sample_mag_summary()`, `plot_mag_mapping()` |
| `nonpareil.R` | `plot_nonpareil_curves()`, `plot_nonpareil_metrics()` |
| `mobile_elements.R` | `mobile_element_summary()`, `plot_mobile_prevalence()` |
| `genomespot.R` | `community_weighted_traits()`, `plot_genomespot_traits()` |

To change a figure, find its function in the script, look up its help page, and add the argument to the call.
To make a figure the templates do not provide, add a script that starts with `gmaio::load_processed()` and uses the same building blocks.
