---
title: "Heatmaps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Heatmaps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
# Show each heatmap exactly as save_plot() writes it, at its automatic size
show_heatmap <- function(heatmap.ht, ...){
  path.s <- knitr::fig_path(".png")
  gmaio::save_plot(heatmap.ht, path.s, dpi = 110, ...)
  knitr::include_graphics(path.s, dpi = 110)
}
```

`plot_heatmap()` draws any profile as a heatmap: features in rows, samples in columns.
This vignette builds up the heatmaps made by `code/heatmaps.R` step by step, then goes through the options for changing them.
It uses the synthetic example project; see `vignette("gmaio", package = "gmaio")` for how projects and profiles work.

```{r project}
project.l <- gmaio::example_project()
metadata.df <- gmaio::analysis_metadata(project.l)
```

## Choosing the data

`plot_heatmap()` draws every feature in the profile it is given, so choosing what to show happens before the call.
`code/heatmaps.R` does the following for each profile and rank:

```{r data}
genus.p <- gmaio::get_profile(project.l, "sylph_taxonomic") |>
  gmaio::aggregate_profile(rank = "genus") |>
  gmaio::filter_profile(min_abundance = 0.5) |>   # at least 0.5% in one sample
  gmaio::top_features(25)                         # then the 25 most abundant by mean
genus.p
```

## The default heatmap

With only a profile and metadata, values are log10 transformed, rows are clustered and samples follow the metadata order (`sample_order` in `config.yml`).
Zeros are drawn in light grey with their own legend key, so they are never confused with small values.

```{r default}
heatmap.ht <- gmaio::plot_heatmap(genus.p, metadata.df, project.l$palettes)
```

```{r default-show, echo = FALSE}
show_heatmap(heatmap.ht)
```

## The template heatmap

The template adds a split by the primary group, a phylum row annotation, axis titles and a legend title:

```{r template}
heatmap.ht <- gmaio::plot_heatmap(genus.p, metadata.df, project.l$palettes,
                                  transform = "log10",
                                  column_split = "Treatment",
                                  row_annotation = "Phylum",
                                  row_title = "Genus",
                                  column_title = "Sample",
                                  legend_title = "Relative\nabundance (%)")
```

```{r template-show, echo = FALSE}
show_heatmap(heatmap.ht)
```

Everything below is a change to this call.
In a project, edit the `gmaio::plot_heatmap()` call in `code/heatmaps.R` and rerun the script.

## Annotations and splits

`column_annotations` adds metadata columns above the heatmap.
Discrete columns use the project colours; numeric columns such as `Age_weeks` get a continuous scale.
`row_split` separates the rows into blocks, here by phylum.

```{r annotations}
heatmap.ht <- gmaio::plot_heatmap(genus.p, metadata.df, project.l$palettes,
                                  column_split = "Treatment",
                                  column_annotations = c("Time", "Age_weeks"),
                                  row_split = "Phylum",
                                  row_labels = "Genus",
                                  row_title = "Genus", column_title = "Sample",
                                  legend_title = "Relative\nabundance (%)")
```

```{r annotations-show, echo = FALSE}
show_heatmap(heatmap.ht)
```

`row_labels = "Genus"` labels rows by the genus column instead of the default `Label` (which includes the phylum).
Annotation colours come from `project.l$palettes`; to override them for one figure use `annotation_colours = list(Time = c(Week_0 = "grey80", Week_4 = "grey30"))`.
To change them everywhere, edit `palettes.yml` or set `colours:` in `config.yml`.

## Values and colours

### Transform

`transform` is one of `"log10"` (default), `"sqrt"`, `"zscore"` or `"none"`.
For log10, half the smallest non-zero value is added before the transform, and the legend shows the original units.
`"zscore"` scales each feature (row) across samples, which suits function profiles where features differ in scale:

```{r zscore}
heatmap.ht <- gmaio::plot_heatmap(genus.p, metadata.df, project.l$palettes, transform = "zscore",
                                  column_split = "Treatment", row_labels = "Genus",
                                  legend_title = "Z-score\n(log10 abundance)")
```

```{r zscore-show, echo = FALSE}
show_heatmap(heatmap.ht)
```

### Colour ramp

`colours` takes a vector of colours (low to high) or the name of any palette in `grDevices::hcl.pals()`, such as `"Mako"`, `"Rocket"`, `"YlOrRd"` or `"Blues 3"`.
Sequential palettes run from light (low) to dark (high).

### Legend breaks and capping the scale

`legend_breaks` sets the legend (and colour) breaks in the original units.
The colours are spread evenly over the breaks, and values above the last break take the last colour, so the top label can say so:

```{r breaks}
heatmap.ht <- gmaio::plot_heatmap(genus.p, metadata.df, project.l$palettes,
                                  column_split = "Treatment", row_labels = "Genus",
                                  colours = "Rocket",
                                  legend_breaks = c(0.01, 0.1, 1, 10, 30),
                                  legend_labels = c("0.01", "0.1", "1", "10", ">= 30"),
                                  legend_title = "Relative\nabundance (%)",
                                  zero_colour = "white", zero_legend = "Not detected")
```

```{r breaks-show, echo = FALSE}
show_heatmap(heatmap.ht)
```

`zero_colour` sets the colour of zero cells (`NULL` colours them from the ramp) and `zero_legend` the label of their legend key (`NULL` for no key).

## Order

Rows are clustered by default (`cluster_rows = TRUE`), within row splits.
With `cluster_rows = FALSE`, `order_rows_by` sets the order: `"abundance"` (mean, highest first), `"label"`, or feature columns such as `"Phylum"` (then by abundance).
Columns follow the metadata order unless `cluster_columns = TRUE`; `order_columns_by` sorts them by `"label"` or metadata columns.

```{r order}
heatmap.ht <- gmaio::plot_heatmap(genus.p, metadata.df, project.l$palettes,
                                  cluster_rows = FALSE, order_rows_by = c("Phylum", "abundance"),
                                  order_columns_by = c("Time", "Treatment"),
                                  column_annotations = c("Time", "Treatment"),
                                  row_annotation = "Phylum", row_labels = "Genus",
                                  legend_title = "Relative\nabundance (%)")
```

```{r order-show, echo = FALSE}
show_heatmap(heatmap.ht)
```

When clustering, `show_row_dend` and `show_column_dend` draw the dendrograms, and `clustering_distance` and `clustering_method` choose the distance (`"euclidean"`, `"manhattan"`, ...) and linkage (`"average"`, `"complete"`, `"ward.D2"`, ...).

## Cell values

`show_values = TRUE` prints each value (in original units) in its cell.
The text is black or white depending on the cell colour, unless `value_colour` is set.
`value_digits` sets the significant digits and `value_size` the font size; zeros are left empty unless `hide_zero_values = FALSE`.
Larger cells make room for the numbers:

```{r values}
top.p <- gmaio::top_features(genus.p, 12)
heatmap.ht <- gmaio::plot_heatmap(top.p, metadata.df, project.l$palettes,
                                  column_split = "Treatment", row_labels = "Genus",
                                  show_values = TRUE, value_digits = 2, value_size = 6,
                                  cell_width = 0.9, cell_height = 0.6,
                                  legend_title = "Relative\nabundance (%)")
```

```{r values-show, echo = FALSE}
show_heatmap(heatmap.ht)
```

## Size, fonts and layout

| Argument | Controls | Default |
|---|---|---|
| `cell_size`, `cell_width`, `cell_height` | Cell size in cm | 0.4 |
| `cell_border_colour`, `cell_border_width` | Lines between cells | white, 0.5 |
| `row_names_size`, `column_names_size` | Row and column label font size | 7 |
| `column_names_rot` | Column label rotation | 90 |
| `row_names_side`, `column_names_side` | Label sides | right, bottom |
| `title_size` | Axis and split titles | 8 |
| `annotation_name_size`, `annotation_size` | Annotation names and track size (cm) | 8, 0.35 |
| `legend_title_size`, `legend_label_size` | Legend fonts | 8, 7 |
| `column_annotation_side`, `row_annotation_side` | Annotation sides | top, left |
| `legend_side`, `merge_legends` | Legend position; one column of legends | right, `FALSE` |
| `padding` | Space around the figure, mm (bottom, left, top, right) | 2, 2, 2, 2 |

The saved figure is sized to fit, so changing cell or font sizes needs no change to `save_plot()`.

```{r layout}
heatmap.ht <- gmaio::plot_heatmap(genus.p, metadata.df, project.l$palettes,
                                  column_split = "Treatment", row_annotation = "Phylum", row_labels = "Genus",
                                  cell_width = 0.3, cell_height = 0.3, column_names_rot = 60,
                                  row_names_size = 6, column_names_size = 6,
                                  legend_side = "bottom", legend_title = "Relative abundance (%)")
```

```{r layout-show, echo = FALSE}
show_heatmap(heatmap.ht)
```

## Anything else

`plot_heatmap()` returns a ComplexHeatmap `Heatmap`.
Any other `ComplexHeatmap::Heatmap()` argument can be passed directly and overrides the gmaio setting, for example wider gaps between splits and a border around each block:

```{r extra}
heatmap.ht <- gmaio::plot_heatmap(genus.p, metadata.df, project.l$palettes,
                                  column_split = "Treatment", row_split = "Phylum", row_labels = "Genus",
                                  column_gap = grid::unit(3, "mm"), border = "grey40")
```

```{r extra-show, echo = FALSE}
show_heatmap(heatmap.ht)
```

Settings applied when the heatmap is drawn (axis titles, legend sides, padding) are stored with it.
`save_plot()` and `draw_heatmap()` apply them, and take any `ComplexHeatmap::draw()` argument to override them, e.g. `gmaio::save_plot(heatmap.ht, path, merge_legend = TRUE)`.

## Saving and viewing

```r
gmaio::save_plot(heatmap.ht, gmaio::figure_path(project.l$config, "heatmaps", "sylph_genus"))
```

The size is measured from the drawn heatmap; pass `width` and `height` (cm) to fix it instead.
To look at a heatmap in RStudio, use `gmaio::draw_heatmap(heatmap.ht)`.
Printing the object also draws it, but without the stored titles and legend layout.

## Other heatmaps

Presence/absence and genome heatmaps (`plot_presence_heatmap()`, `plot_genome_matrix()`, `plot_dram_heatmap()`) are separate functions for isolate and MAG data; see their help pages.
