---
title: "Isolate genomes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Isolate genomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
# Show each figure exactly as save_plot() writes it, at its automatic size
show_plot <- function(plot, ...){
  path.s <- knitr::fig_path(".png")
  gmaio::save_plot(plot, path.s, dpi = 110, ...)
  knitr::include_graphics(path.s, dpi = 110)
}
has_tree.b <- requireNamespace("ggtree", quietly = TRUE) && requireNamespace("ape", quietly = TRUE)
```

In isolate mode the pipeline assembles, types and annotates each genome, and compares genomes within comparison groups (ANI, cgMLST, pangenome and a tree).
An isolate project is created with `gmaio::create_project(path, mode = "isolate")` and has its own scripts: `main.R`, then one script per analysis.
This vignette goes through them on a synthetic example: 20 *Escherichia coli* isolates from four sequence types and three sources, plus two reference genomes.
See `vignette("gmaio", package = "gmaio")` for how projects, profiles and colours work; they are the same in both modes.

```{r project}
project.l <- gmaio::example_project(mode = "isolate")
project.l
metadata.df <- gmaio::analysis_metadata(project.l)
head(metadata.df)
```

`Source` is the primary group (`analysis: group_variables`), used for colours and tests.

## What main.R loads

| Step | Function | Result |
|---|---|---|
| Quality, taxonomy, annotation, typing | `add_isolate_genomes()` | `tables`: `checkm2`, `gtdbtk`, `bakta`, `mlst`, `amrfinder`, `isescan` and a one-row-per-isolate `genome_summary`; `profiles`: `amr_genes`, `amr_classes`, `is_families` |
| Comparison groups | `add_comparisons()` | `tables$comparisons[[group]]`: `entries`, `pangenome`, `ani`, `cgmlst`, `tree` |
| Mobile elements | `add_mobile_elements()` | `tables$virus_summary` (geNomad) and plasmid tables |
| Traits | `add_genomespot()` | GenomeSPOT predictions, when run |

`write_isolate_tables()` writes them to `Result_tables/`.

## Genome summary

`code/genome_summary.R` draws assembly and quality metrics side by side:

```{r summary}
summary.gg <- gmaio::plot_genome_summary(project.l$tables$genome_summary, metadata.df, fill_by = "Source",
                                         colours.v = project.l$palettes$Source)
```

```{r summary-show, echo = FALSE}
show_plot(summary.gg)
```

Isolates follow the metadata order (`sample_order`).
`metrics` chooses and orders the panels (any of `Genome_size`, `Contigs`, `Contig_N50`, `GC_content`, `Completeness`, `Contamination`, `Mean_coverage`, `CDS`), and `bar_width` and `base_size` change the bars and text.

## Genome metadata and sequence types

Comparison figures cover samples and references, so they take `genome_metadata()` rather than the analysis metadata.
It adds `Entry_type` (Sample or Reference), gives references their accession as label and `"Reference"` for the group variables, and adds each isolate's MLST sequence type (`ST`) and GTDB species:

```{r genome-metadata}
group.l <- project.l$tables$comparisons$all
genomes.df <- gmaio::genome_metadata(project.l, "all")
table(genomes.df$ST, genomes.df$Source)
```

Any of these columns can be used to annotate, split or colour the figures below.

## AMR genes and insertion sequences

`code/amr_mobile_elements.R` draws AMRFinderPlus genes as presence/absence, split by drug class:

```{r amr}
amr.p <- gmaio::get_profile(project.l, "amr_genes")
amr.ht <- gmaio::plot_presence_heatmap(amr.p, genomes.df, project.l$palettes, c("ST", "Source"), row_split = "Class")
```

```{r amr-show, echo = FALSE}
show_plot(amr.ht)
```

`plot_presence_heatmap()` options: `column_split` (e.g. `"ST"`), `colours` (absent and present), `cluster_rows` and `cluster_columns`, `show_row_dend` and `show_column_dend`, `show_row_names`, `label_by`, `cell_width` and `cell_height`, `row_names_size` and `column_names_size`, and `max_features`.

```{r amr-split}
amr.ht <- gmaio::plot_presence_heatmap(amr.p, genomes.df, project.l$palettes, "Source",
                                       row_split = "Class", column_split = "ST", show_row_dend = FALSE,
                                       colours = c(Absent = "grey95", Present = "#B2182B"))
```

```{r amr-split-show, echo = FALSE}
show_plot(amr.ht)
```

`gene_association()` tests each gene against a variable with Fisher's exact test (BH adjusted).
The script tests the primary group; here AMR genes follow the sequence type much more than the source:

```{r association}
by_st.df <- gmaio::gene_association(amr.p, genomes.df, "ST")
utils::head(by_st.df[, c("Feature_ID", "N_present", "Prevalence_ST131", "Prevalence_ST10", "Enriched_in", "Q_value")])
```

Gene counts per drug class (`amr_classes`) and insertion sequences per family (`is_families`) are count profiles, drawn with `plot_heatmap()` and printed values:

```{r is}
is.ht <- gmaio::plot_heatmap(gmaio::get_profile(project.l, "is_families"), metadata.df, project.l$palettes,
                             transform = "none", column_split = "Source", cluster_rows = FALSE, show_values = TRUE,
                             legend_title = "Copies", cell_size = 0.6, colours = c("#F7F7F7", "#6BAED6", "#08306B"))
```

```{r is-show, echo = FALSE}
show_plot(is.ht)
```

All `plot_heatmap()` options are described in `vignette("heatmaps", package = "gmaio")`.

## Average nucleotide identity

`code/ani.R` turns fastANI results into a matrix (both directions averaged) and draws it:

```{r ani}
ani.m <- gmaio::ani_matrix(group.l$ani)
ani.ht <- gmaio::plot_genome_matrix(ani.m, genomes.df, project.l$palettes, c("Entry_type", "ST", "Source"))
```

```{r ani-show, echo = FALSE}
show_plot(ani.ht)
```

Genomes are clustered on 100 - ANI.
Values are printed for up to 25 genomes, sized to fit the cells.

| Argument | Controls | Default |
|---|---|---|
| `split_by` | Split rows and columns into blocks (e.g. `"ST"`), clustering within blocks | none |
| `cluster`, `show_dend` | Cluster genomes, and draw the dendrograms | `TRUE`, `TRUE` |
| `breaks`, `colours` | Colour breaks and the colours at them | 85, 95, 97, 99, 100 in blues |
| `legend_title` | Legend title | ANI (%) |
| `show_values`, `value_digits`, `value_size` | Printed values | up to 25 genomes, 1, fitted |
| `label_by` | Genome labels | `Sample_label` |
| `cell_size`, `names_size` | Cell size (cm) and label font size | 0.45, 7 |
| `row_title`, `column_title` | Axis titles | none |

Genomes of one species share at least about 95% ANI, and these are all between 97 and 100%, so narrower breaks show more detail:

```{r ani-split}
ani.ht <- gmaio::plot_genome_matrix(ani.m, genomes.df, project.l$palettes, "Source", split_by = "ST",
                                    breaks = c(97, 98, 99, 99.5, 100), show_values = FALSE, cell_size = 0.35)
```

```{r ani-split-show, echo = FALSE}
show_plot(ani.ht)
```

The script also ordinates the ANI distances.
`ordinate_distance()` turns any distance matrix into an ordination for `plot_ordination()` and the PERMANOVA functions (see `vignette("ordination", package = "gmaio")`):

```{r ani-pcoa}
ordination <- gmaio::ordinate_distance(100 - ani.m, "100 - ANI")
pcoa.gg <- gmaio::plot_ordination(ordination, genomes.df, "ST", shape_by = "Source", title = "ANI")
```

```{r ani-pcoa-show, echo = FALSE}
show_plot(pcoa.gg, width = 16, height = 13)
```

Isolates of one sequence type sit on top of each other, so labels are left out here; the script adds them with `label_by = "Sample_label"`, which suits more diverse collections.

## cgMLST

`code/cgmlst.R` counts allele differences between genomes from chewBBACA's cgMLST matrix, ignoring loci missing in either genome:

```{r cgmlst}
distance.m <- gmaio::cgmlst_distances(group.l$cgmlst$cgMLST95)
cgmlst.ht <- gmaio::plot_genome_matrix(distance.m, genomes.df, project.l$palettes, c("ST", "Source"), type = "distance")
```

```{r cgmlst-show, echo = FALSE}
show_plot(cgmlst.ht)
```

For distances, breaks default to 0, 5, 10, 25, 50, 100 and 200 alleles in yellow to red; set `breaks` to highlight the threshold you use for transmission clusters, e.g. `breaks = c(0, 10, 50, 300)`.

## Pangenome

`code/pangenome.R` classifies Panaroo gene clusters by prevalence (core in at least 99% of genomes, soft core 95%, shell 15%, cloud below):

```{r categories}
pangenome.p <- group.l$pangenome
categories.gg <- gmaio::plot_pangenome_categories(pangenome.p)
```

```{r categories-show, echo = FALSE}
show_plot(categories.gg)
```

The accessory genes (shell and cloud) are drawn as presence/absence, keeping the `max_features` (150) most variable:

```{r accessory}
accessory.ht <- gmaio::plot_presence_heatmap(pangenome.p, genomes.df, project.l$palettes, c("ST", "Source"),
                                             show_row_names = FALSE, cell_height = 0.08)
```

```{r accessory-show, echo = FALSE}
show_plot(accessory.ht)
```

The script also ordinates the accessory genes (Jaccard PCoA) and tests them against the primary group with `gene_association()`.

## Phylogeny

`code/phylogeny.R` draws the IQ-TREE tree of each comparison group, midpoint rooted, with tips coloured by the primary group:

```{r tree, eval = has_tree.b}
tree_genomes.df <- gmaio::genome_metadata(project.l, "all", group.l$tree$tip.label)
tree.gg <- gmaio::plot_tree(group.l$tree, tree_genomes.df, "Source", project.l$palettes$Source, shape_by = "Entry_type")
```

```{r tree-show, echo = FALSE, eval = has_tree.b}
show_plot(tree.gg)
```

| Argument | Controls | Default |
|---|---|---|
| `colour_by`, `colours.v` | Tip point fill | none |
| `shape_by`, `shapes` | Tip point shape (samples circles and references diamonds for `Entry_type`) | none |
| `label_by`, `label_size` | Tip labels (`NULL` for none) and their size | `Sample_label`, 2.4 |
| `align_labels` | Line labels up at the right with dotted leaders | `FALSE` |
| `point_size` | Tip point size | 2 |
| `layout` | `"rectangular"` or `"circular"` | rectangular |
| `midpoint`, `ladderize` | Midpoint root (needs phangorn), ladderize | `TRUE`, `TRUE` |
| `scale_bar`, `x_label` | Scale bar instead of an axis, and the axis title | `FALSE`, substitutions per site |
| `label_space` | Room for labels, as a fraction of the tree width | from the longest label |
| `width`, `height` | Figure size (cm) | from the number of tips |

```{r tree-options, eval = has_tree.b}
tree.gg <- gmaio::plot_tree(group.l$tree, tree_genomes.df, "ST", shape_by = "Source", align_labels = TRUE,
                            scale_bar = TRUE)
```

```{r tree-options-show, echo = FALSE, eval = has_tree.b}
show_plot(tree.gg)
```

The result is a ggtree plot, so ggtree layers can be added, e.g. `tree.gg + ggtree::geom_nodelab(size = 2)` to show node support.

## Prophages and plasmids

geNomad results are summarised by virus taxonomy per isolate:

```{r viruses}
virus.gg <- gmaio::plot_virus_taxonomy(project.l$tables$virus_summary, metadata.df, facet_variable = "Source")
```

```{r viruses-show, echo = FALSE}
show_plot(virus.gg)
```

## Saving

Every figure is saved with `save_plot()`, sized automatically:

```r
gmaio::save_plot(ani.ht, gmaio::figure_path(project.l$config, "ani", "all__ani_heatmap"))
```
