---
title: "Output files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Output files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# Names as code (so <Set> and <dataset> placeholders show), other text with < and > escaped
show_table <- function(x.df){
  for (column.s in names(x.df)){
    values.v <- as.character(x.df[[column.s]])
    if (column.s == "Written_by") values.v <- sub("^code/", "", values.v)
    x.df[[column.s]] <- if (column.s %in% c("File", "Sheet", "Folder", "Column", "Set", "Written_by")){
      paste0("`", values.v, "`")
    } else if (column.s == "Files"){
      vapply(strsplit(values.v, ", ", fixed = TRUE), function(x) paste0("`", x, "`", collapse = ", "), character(1))
    } else {
      gsub("(--[a-z_]+)", "`\\1`", gsub(">", "&gt;", gsub("<", "&lt;", values.v)))
    }
  }
  knitr::kable(x.df, row.names = FALSE)
}
```

A gmaio project writes three folders, set in `outputs:` of `config.yml`:

| Folder | Contents |
|---|---|
| `Result_tables/` | Excel tables: the data from `main.R`, and the statistics from each analysis script |
| `Result_figures/` | One subfolder of figures per analysis script |
| `Result_other/` | `processed.rds` (the processed project the scripts load) and the working files of some methods (e.g. `maaslin3/`) |

This vignette explains what is in each file, and in particular the MAG files, which come in one version per genome set.
You do not need it open while working: the same descriptions travel with the files.

- Every workbook gmaio writes has an **`About`** sheet first, describing the file, each of its sheets and the columns that need explaining (and, for MAG files, which genome set it is).
- `Result_tables/README.md` and `Result_figures/README.md` list every file or folder present with what it contains.
  They are rewritten whenever gmaio writes a table or saves a figure, so they match the files there.

In R, `output_catalogue()`, `figure_catalogue()`, `mag_set_catalogue()` and `describe_output()` return the same information.

## MAG genome sets

The pipeline maps every sample's reads to up to six sets of genomes.
Each set gives its own `MAG_<set>_abundances.xlsx` and `MAG_<set>_taxonomy_relative_abundance.xlsx`, and its own `Reads_mapped_<label>` columns in the read statistics.
The sets differ in two ways: which bins they keep (any quality, or high quality only) and whether genomes are pooled across samples.

```
All bins, all samples ──┬── dereplicate at 95% ANI ──────────────────────────► derep_bins
                        │        └── keep the representatives that are HQ ────► hq_bins
                        ├── keep HQ bins ── dereplicate at 95% ANI ──────────► hq_derep_bins
                        └── keep HQ bins + reference genomes ── dereplicate ──► hq_ref_bins

Each sample's own bins ─┬── dereplicate within the sample ───────────────────► ws_derep_bins
                        └── keep HQ bins ── dereplicate within the sample ───► ws_hq_bins
```

HQ means completeness - 3 x contamination >= 50, in CheckM2 or CheckM1 (pipeline `--hq_quality_source`).

```{r sets, echo = FALSE}
sets.df <- gmaio::mag_set_catalogue()
show_table(sets.df[, c("Set", "Name", "Genomes", "Use_for")])
```

The `Name` column is how each set is labelled in figures, e.g. in `reads_mapped_to_mags`.

### Which one to use

- **To compare samples** (barcharts, heatmaps, ordination, differential abundance), use **`hq_derep_bins`**.
  Its genomes are pooled across samples and all high quality, so a genome absent from a sample really was not detected there.
  The templates use it.
- **`hq_bins` vs `hq_derep_bins`**: both hold HQ genomes pooled across samples, but they are built in opposite orders.
  `hq_bins` dereplicates all bins first and keeps the representatives that happen to be HQ, so a cluster whose chosen representative is a poorer bin is lost even when it contains an HQ bin.
  `hq_derep_bins` keeps the HQ bins first and then dereplicates them, so every HQ genome is represented.
  `hq_derep_bins` therefore has at least as many genomes; often the two are nearly the same, but they differ when HQ genomes cluster with poorer bins.
- **`derep_bins`** keeps genomes of any quality: the most complete picture of what was assembled, and the most reads explained, but low-quality genomes have unreliable taxonomy and functions.
- **`ws_derep_bins` and `ws_hq_bins`** (only with the pipeline's `--within_sample_dereplication`) map each sample only to the genomes assembled from it.
  They answer "how much of this sample do its own MAGs explain?", shown in `MAG_per_sample_recovery.xlsx` and the `reads_mapped_to_mags` figure.
  A genome can only be found in the sample it came from, so these sets say nothing about differences between samples; gmaio warns if you ordinate or test them.

### The four measures in each abundance file

| Sheet | What it is | Sums to |
|---|---|---|
| `relative_abundance` | Mean coverage of each genome as a share of all genomes in the set: the share of genome copies | 100 per sample |
| `coverm_relative_abundance` | CoverM relative abundance: % of the sample's reads (after QC and host removal) that map to the genome | the % of reads mapped to the set |
| `coverage` | Mean read depth over the genome (x) | - |
| `read_count` | Reads mapped to the genome | reads mapped |

Use `relative_abundance` to compare composition between samples; it is what the templates plot and test.
Use `coverm_relative_abundance` when the unmapped share matters, e.g. "this genome is 5% of all the reads in the sample".
`MAG_<set>_taxonomy_relative_abundance.xlsx` is `relative_abundance` summed to each GTDB rank.

### Finding a MAG across sets

`MAG_summary.xlsx` has every bin from every sample, with its quality and taxonomy.
For each set it has two columns: `Is_representative_<set>` (the bin is in that set) and `Representative_<set>` (the bin that represents its cluster there).
So to follow one genome from its sample into the pooled catalogue:

```r
bins.df <- readxl::read_excel("Result_tables/MAG_summary.xlsx", "Bins")
bins.df[bins.df$Bin_ID == "<a bin>", c("High_quality", "Representative_hq_derep_bins")]
```

In the abundance tables, the first column (`genome`) is the `Bin_ID` of each representative, and `Short_ID` (`MAG001`, ...) is the short name used in figures.

## Tables

The files `main.R` writes (the data) and those the analysis scripts write (the statistics):

```{r tables-metagenome, echo = FALSE}
show_table(gmaio::output_catalogue("metagenome")[, c("File", "Written_by", "Description")])
```

Isolate projects write:

```{r tables-isolate, echo = FALSE}
isolate.df <- gmaio::output_catalogue("isolate")
show_table(isolate.df[isolate.df$Mode == "isolate" | isolate.df$File == "Metadata_and_read_stats.xlsx",
                      c("File", "Written_by", "Description")])
```

`describe_output()` gives the sheets and key columns of any of them; this is what the `About` sheet shows:

```{r describe}
about.l <- gmaio::describe_output("MAG_summary.xlsx")
show_table(about.l$Sheets[, c("Sheet", "Description")])
show_table(utils::head(about.l$Columns[, c("Column", "Description")], 8))
```

Taxonomic tables (sylph, SingleM and the MAG taxonomy files) have one sheet per rank, from phylum to species.
Rows are taxa, named by their full GTDB lineage in the first column with a short `Label`, and columns are samples by `Sample_ID`.
The `Metadata` sheet of `Metadata_and_read_stats.xlsx` links each `Sample_ID` to its label and metadata.

## Figures

Each analysis script saves to its own folder of `Result_figures/`.
`<dataset>` is a profile and rank, such as `sylph_taxonomic__genus`, or a MAG profile such as `mags_hq_derep_bins_relative_abundance`.

```{r figures, echo = FALSE}
show_table(gmaio::figure_catalogue()[, c("Folder", "Written_by", "Files", "Description")])
```

To change a figure, edit the plot function call in the script that writes it; the other vignettes explain the options.
