---
title: "Strain sharing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Strain sharing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
# Show each figure exactly as save_plot() writes it, at its automatic size
show_plot <- function(plot, ...){
  path.s <- knitr::fig_path(".png")
  gmaio::save_plot(plot, path.s, dpi = 110, ...)
  knitr::include_graphics(path.s, dpi = 110)
}
```

With `--run_instrain` and `--run_tracs`, the pipeline compares the strains of high-quality bins between samples.
inStrain profiles the reads of each sample against the bins and compares samples by popANI, calling the same strain at popANI >= 99.999%.
TRACS gives SNP distances between samples for each genome.
`code/strains.R` asks which samples share strains, whether strains are shared more within groups than between them, and how diverse each sample's strain populations are.
This vignette follows that script on the synthetic example project, where mice sharing a cage usually share strains and treated mice have less diverse strains at week 4.

```{r project}
project.l <- gmaio::example_project()
metadata.df <- gmaio::analysis_metadata(project.l)
strains.l <- project.l$tables$strains
names(strains.l)
```

`main.R` loads the results with `add_strains()`.
inStrain names samples by their BAM files (`S01.bam`); gmaio links them to the metadata like any other output, and adds each genome's short ID and label from the bin summary.

```{r sharing}
head(strains.l$instrain_sharing[, c("Short_ID", "Sample_ID_a", "Sample_ID_b", "popANI", "Same_strain")])
```

## Who shares strains with whom

`strain_sharing_matrix()` turns pairs of samples into a sample by sample matrix: here the number of genomes each pair shares as the same strain, from inStrain's counts.

```{r counts}
counts.df <- strains.l$instrain_counts
samples.v <- intersect(metadata.df$Sample_ID, c(counts.df$Sample_ID_a, counts.df$Sample_ID_b))
counts.m <- gmaio::strain_sharing_matrix(counts.df, "Same_strain", samples.v, missing = 0)
counts.ht <- gmaio::plot_strain_sharing_matrix(counts.m, metadata.df, project.l$palettes, group = "Cage")
```

```{r counts-show, echo = FALSE}
show_plot(counts.ht)
```

For one genome, the popANI between every pair of samples it was compared in shows how close the strains are; grey pairs were not compared (too little coverage in one of them).

```{r popani}
genome.df <- strains.l$instrain_sharing[strains.l$instrain_sharing$Short_ID == "Bin001", ]
genome.df$popANI_percent <- 100 * genome.df$popANI
ani.m <- gmaio::strain_sharing_matrix(genome.df, "popANI_percent", samples.v, diagonal = 100)
ani.ht <- gmaio::plot_strain_sharing_matrix(ani.m, metadata.df, project.l$palettes, group = "Cage", type = "popani",
                                            title = genome.df$Label[1])
```

```{r popani-show, echo = FALSE}
show_plot(ani.ht)
```

## Within and between groups

`compare_pairs_by_group()` splits sample pairs into pairs from the same group and pairs from different groups, and tests the difference with a permutation test that shuffles group labels between samples.
Shuffling samples, not pairs, keeps together all the genomes a pair of samples shares, so the test does not treat them as independent.
It works for strain sharing (a logical value: the percent of pairs sharing a strain) and for distances such as TRACS SNP distances.

```{r within-between}
shared.l <- gmaio::compare_pairs_by_group(strains.l$instrain_sharing, "Same_strain", metadata.df, "Cage",
                                          permutations = 999)
shared.l$summary
shared.l$test
sharing.gg <- gmaio::plot_pairs_by_group(shared.l, project.l$palettes$Cage)
```

```{r within-between-show, echo = FALSE}
show_plot(sharing.gg)
```

```{r tracs}
distances.l <- gmaio::compare_pairs_by_group(strains.l$tracs_distances, "Filtered_SNP_distance", metadata.df, "Cage",
                                             permutations = 999)
tracs.gg <- gmaio::plot_pairs_by_group(distances.l, project.l$palettes$Cage, y_label = "TRACS SNP distance (filtered)")
```

```{r tracs-show, echo = FALSE}
show_plot(tracs.gg)
```

## Strain diversity

inStrain reports nucleotide diversity for each genome in each sample.
Testing genome by genome would count a sample once per genome, so `instrain_sample_summary()` first summarises each sample over its well-covered genomes (`min_breadth`: the share of the genome inStrain covers at its minimum depth).

```{r diversity}
diversity.df <- gmaio::instrain_sample_summary(strains.l$instrain_genomes, min_breadth = 0.5)
metadata.df$Treatment_time <- factor(paste(metadata.df$Treatment, metadata.df$Time),
                                     levels = c("Control Week_0", "Control Week_4", "Treated Week_0", "Treated Week_4"))
diversity.l <- gmaio::plot_instrain_diversity(diversity.df, metadata.df, "Treatment_time")
```

```{r diversity-show, echo = FALSE}
show_plot(diversity.l$plot)
```
