---
title: "Strain sharing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Strain sharing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
# Show each figure exactly as save_plot() writes it, at its automatic size
show_plot <- function(plot, ...){
  path.s <- knitr::fig_path(".png")
  gmaio::save_plot(plot, path.s, dpi = 110, ...)
  knitr::include_graphics(path.s, dpi = 110)
}
```

With `--run_instrain` and `--run_tracs`, the pipeline compares the strains of high-quality bins between samples.
inStrain profiles the reads of each sample against the bins and compares samples by popANI, calling the same strain at popANI >= 99.999%.
TRACS gives SNP distances between samples for each genome.
`code/strains.R` asks which samples share strains, whether strains are shared more within groups than between them, and how diverse each sample's strain populations are.
This vignette follows that script on the synthetic example project, where mice sharing a cage usually share strains and treated mice have less diverse strains at week 4.

```{r project}
project.l <- gmaio::example_project()
metadata.df <- gmaio::analysis_metadata(project.l)
strains.l <- project.l$tables$strains
names(strains.l)
```

`main.R` loads the results with `add_strains()`.
inStrain names samples by their BAM files (`S01.bam`); gmaio links them to the metadata like any other output, and adds each genome's short ID and label from the bin summary, or from the reference genomes when the strain set includes them (`--strain_genome_source hq_ref_representatives` or `references`).
`strain_genomes` lists every genome the pipeline considered, whether it was kept and its `origin` (`mag` or `reference`).

```{r sharing}
head(strains.l$instrain_sharing[, c("Short_ID", "Sample_ID_a", "Sample_ID_b", "popANI", "Same_strain")])
```

## Who shares strains with whom

`strain_sharing_matrix()` turns pairs of samples into a sample by sample matrix: here the number of genomes each pair shares as the same strain, from inStrain's counts.

```{r counts}
counts.df <- strains.l$instrain_counts
samples.v <- intersect(metadata.df$Sample_ID, c(counts.df$Sample_ID_a, counts.df$Sample_ID_b))
counts.m <- gmaio::strain_sharing_matrix(counts.df, "Same_strain", samples.v, missing = 0)
counts.ht <- gmaio::plot_strain_sharing_matrix(counts.m, metadata.df, project.l$palettes, group = "Cage")
```

```{r counts-show, echo = FALSE}
show_plot(counts.ht)
```

For one genome, the popANI between every pair of samples it was compared in shows how close the strains are; grey pairs were not compared (too little coverage in one of them).

```{r popani}
genome.df <- strains.l$instrain_sharing[strains.l$instrain_sharing$Short_ID == "Bin001", ]
genome.df$popANI_percent <- 100 * genome.df$popANI
ani.m <- gmaio::strain_sharing_matrix(genome.df, "popANI_percent", samples.v, diagonal = 100)
ani.ht <- gmaio::plot_strain_sharing_matrix(ani.m, metadata.df, project.l$palettes, group = "Cage", type = "popani",
                                            title = genome.df$Label[1])
```

```{r popani-show, echo = FALSE}
show_plot(ani.ht)
```

## Within and between groups

`compare_pairs_by_group()` splits sample pairs into pairs from the same group and pairs from different groups, and tests the difference with a permutation test that shuffles group labels between samples.
Shuffling samples, not pairs, keeps together all the genomes a pair of samples shares, so the test does not treat them as independent.
It works for strain sharing (a logical value: the percent of pairs sharing a strain) and for distances such as TRACS SNP distances.

```{r within-between}
shared.l <- gmaio::compare_pairs_by_group(strains.l$instrain_sharing, "Same_strain", metadata.df, "Cage",
                                          permutations = 999)
shared.l$summary
shared.l$test
sharing.gg <- gmaio::plot_pairs_by_group(shared.l, project.l$palettes$Cage)
```

```{r within-between-show, echo = FALSE}
show_plot(sharing.gg)
```

```{r tracs}
distances.l <- gmaio::compare_pairs_by_group(strains.l$tracs_distances, "Filtered_SNP_distance", metadata.df, "Cage",
                                             permutations = 999)
tracs.gg <- gmaio::plot_pairs_by_group(distances.l, project.l$palettes$Cage, y_label = "TRACS SNP distance (filtered)")
```

```{r tracs-show, echo = FALSE}
show_plot(tracs.gg)
```

## Strain diversity

inStrain reports nucleotide diversity for each genome in each sample.
Testing genome by genome would count a sample once per genome, so `instrain_sample_summary()` first summarises each sample over its well-covered genomes (`min_breadth`: the share of the genome inStrain covers at its minimum depth).

As for alpha diversity, `show_test = FALSE` and `brackets = FALSE` leave the statistics off the figure (or `show_statistics.b <- FALSE` at the top of `code/strains.R`).

```{r diversity}
diversity.df <- gmaio::instrain_sample_summary(strains.l$instrain_genomes, min_breadth = 0.5)
metadata.df$Treatment_time <- factor(paste(metadata.df$Treatment, metadata.df$Time),
                                     levels = c("Control Week_0", "Control Week_4", "Treated Week_0", "Treated Week_4"))
diversity.l <- gmaio::plot_instrain_diversity(diversity.df, metadata.df, "Treatment_time")
```

```{r diversity-show, echo = FALSE}
show_plot(diversity.l$plot)
```

## Comparison samples

With `--strain_include_comparison_reads true`, the pipeline also profiles the comparison samples (`--comparison_reads`) against the same genomes and compares them with the study's samples, so the tables answer whether a study sample carries the same strain as a sample from another study.
`add_strains()` links them to the comparison metadata, so `main.R` runs `add_comparison()` first; without comparison metadata they are left out, with a message.
Every table says which dataset each sample belongs to (`Dataset_a` and `Dataset_b` for pairs, `Dataset` otherwise; the labels are `comparison: dataset_labels` in `config.yml`), and `samples` lists the samples of each dataset.
The analyses above use the study's samples only.

```{r comparison-samples}
table(strains.l$samples$Dataset)
```

`combined_metadata()` puts the two datasets' samples together, with the study's primary group and the comparison group variable in one `Comparison_group` column.
`cross_dataset_pairs()` keeps the pairs between a study sample and a comparison sample, study sample first, with both samples' labels and groups: here, the study mice that share a strain with a mouse from another facility.

```{r cross-dataset}
combined.df <- gmaio::combined_metadata(project.l)
shared.df <- strains.l$instrain_sharing[strains.l$instrain_sharing$Same_strain, ]
cross.df <- gmaio::cross_dataset_pairs(shared.df, combined.df)
head(cross.df[, c("Short_ID", "Sample_label_a", "Group_a", "Sample_label_b", "Group_b", "popANI")])
```

`strain_sharing_by_genome()` counts, for each genome, the pairs compared and those with the same strain within the study, between the datasets and within the comparison samples.

```{r by-genome}
genomes.df <- gmaio::strain_sharing_by_genome(strains.l$instrain_sharing, strains.l$strain_genomes,
                                              levels(combined.df$Dataset), combined.df$Sample_ID)
head(genomes.df[, -(1:4)], 3)
```

`compare_pairs_by_group()` summarises pairs for every pair of groups across both datasets.
Groups from different datasets also differ in how the samples were collected and sequenced, so `test = FALSE` leaves out the within versus between groups test, which would pool both kinds of difference.
`plot_group_pair_matrix()` shows the percent of pairs sharing a strain for every pair of groups, with the number of pairs compared, split into blocks by dataset.
Here mice of one cage group or facility often share strains, and study mice share strains with Facility_A more than with Facility_B.

```{r group-pairs}
by_group.l <- gmaio::compare_pairs_by_group(strains.l$instrain_sharing, "Same_strain", combined.df, "Comparison_group",
                                            test = FALSE)
group_datasets.v <- stats::setNames(as.character(combined.df$Dataset), combined.df$Comparison_group)
group_datasets.v <- group_datasets.v[!duplicated(names(group_datasets.v))]
groups.gg <- gmaio::plot_group_pair_matrix(by_group.l, datasets.v = group_datasets.v)
```

```{r group-pairs-show, echo = FALSE}
show_plot(groups.gg)
```

The same works for TRACS SNP distances (the median for each pair of groups) and for the sample by sample matrix of shared strains (`plot_strain_sharing_matrix()` with `combined.df` and `group = "Comparison_group"`).
`code/strains.R` writes these to `Strain_sharing.xlsx` (`Shared_with_comparison`, `Sharing_by_dataset_pair`, `Comparison_sharing_by_groups`, `Comparison_TRACS_by_groups`, `Comparison_diversity_tests`) and the `*_with_comparison` figures.

## Samples and genomes behind each output

Every output uses the genomes the pipeline compared strains on, set by `--strain_genome_source`: HQ MAGs only (`hq_representatives`, the default), HQ MAGs and reference genomes dereplicated together (`hq_ref_representatives`), or the reference genomes only (`references`).
`Strain_genomes` lists them with their `origin`, and `Genome_sharing` gives the counts per genome.
The outputs differ only in their samples:

| Output | Samples |
|---|---|
| `shared_strains_per_sample_pair`, `popani__<Short_ID>`, `strain_sharing_within_between_groups`, `tracs_snp_distance_within_between_groups`, `strain_diversity_by_group`; sheets `Sharing_by_*`, `Sharing_test`, `TRACS_by_*`, `TRACS_test`, `Sample_diversity_tests`, `Sample_diversity_pairwise` | The study's analysis samples |
| `shared_strains_per_sample_pair_with_comparison`, `strain_sharing_group_pairs_with_comparison`, `tracs_snp_distance_group_pairs_with_comparison`, `strain_diversity_by_group_with_comparison`; sheets `Shared_with_comparison`, `Sharing_by_dataset_pair`, `Comparison_*` | The study's and the comparison samples (`--strain_include_comparison_reads`) |
| Sheets `inStrain_*`, `TRACS_distances`, `TRACS_clusters`, `Sample_diversity` | Every sample in the pipeline results, with its `Dataset` |
| Sheet `Genome_sharing` | The study's and the comparison samples, without excluded samples |

inStrain and TRACS compare samples two at a time, so including the comparison reads does not change the results for pairs of study samples.
